### 2021-12-06 paper.tex
initial commit

\documentclass{scrartcl}

\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath}
\usepackage[german]{authblk}
\usepackage{hyperref}

\newcommand{\mytitle}{Notes on the high-performance implementation of special functions}
\newcommand{\myauthor}{Alexander Voigt}

\hypersetup{
  pdftitle   = {\mytitle},
  pdfauthor  = {\myauthor},
  colorlinks = true,
  linkcolor  = blue,
  citecolor  = blue,
  urlcolor   = blue,
  linkcolor  = blue
}

\title{\mytitle}
\author{\myauthor}
\date{\today}

\begin{document}
\maketitle

\section*{Abstract}

\section{Introduction}

\cite{polylogarithm}

\section{Comparison of algorithms}

\section{Conclusions}

\bibliographystyle{ieeetr}
\bibliography{paper}

\end{document}

### 2021-12-06 paper.tex
extend text

@@
-\documentclass{scrartcl}
+\documentclass[10pt,DIV16,twocolumn]{scrartcl}
 
@@
 \usepackage[german]{authblk}
+\usepackage[format=plain,labelfont={bf}]{caption}
+\usepackage{listings}
 \usepackage{hyperref}
@@
 \newcommand{\myauthor}{Alexander Voigt}
+
+% source code highlighting
+\lstset{breaklines=true,
+  breakatwhitespace=true,
+  % numbers=left,
+  % stepnumber=1,
+  basicstyle=\ttfamily\lst@ifdisplaystyle\footnotesize\fi, % code block in smaller font
+  commentstyle=\ttfamily,
+  prebreak={\textbackslash},
+  breakindent=10pt,
+  breakautoindent=false,
+  showspaces=false,
+  showstringspaces=false,
+  frame=shadowbox,
+  rulesep=0.1em,
+  abovecaptionskip=0em,
+  aboveskip=1.5em,
+  belowcaptionskip=0.5em,
+  belowskip=1em,
+  keywordstyle=\color{blue},
+  stringstyle=\color{red},
+  commentstyle=\color{green},
+}
 
@@
 \author{\myauthor}
+\affil{Fachbereich Energie und Biotechnologie, Hochschule Flensburg, Kanzleistraße 91--93, 24943 Flensburg, Germany}
 \date{\today}
@@
 
+As an example we consider the function $\log(1+x)$, which has the
+following Taylor expansion around $x=0$ for $-1 < x \leq 1$:
+%
+\begin{align}
+  \log(1+x) \approx x - \frac{x^2}{2} + \frac{x^3}{3} - \frac{x^4}{4} + \frac{x^5}{5} - \frac{x^6}{6} + \frac{x^7}{7},
+\end{align}
+%
+where terms of $O(x^8)$ have been omitted.
+In C this function could be implemented as follows:
+%
+\begin{lstlisting}[language=C]
+double log1p_naive(double x) {
+   return x - x*x/2 + x*x*x/3 - x*x*x*x/4
+      + x*x*x*x*x/5 - x*x*x*x*x*x/6
+      + x*x*x*x*x*x*x/7;
+}
+\end{lstlisting}
+%
+This naive implementation requires 26 floating-point multiplications
+and 6 floating-point additions.\footnote{Modern compilers usually
+  rewrite expensive floating-point divisions by a constant, such as
+  $x/2$, as less expensive floating-point multiplication by a
+  constant, for example as $0.5x$.}
+
+As is well known, the performante of this implementation can be
+improved by applying Horner's scheme, which reduces the number of
+expensive floating-point multiplications to 7 and requires 6
+additions:
+%
+\begin{align}
+  \log(1+x) \approx{}& x \Bigg\{1 + x \Bigg[-\frac{1}{2} + x \Bigg(\frac{1}{3} + x \Bigg\{-\frac{1}{4}  \notag \\
+  &+ x \Bigg[\frac{1}{5} + x\Bigg(-\frac{1}{6} + \frac{x}{7}\Bigg)\Bigg]\Bigg\}\Bigg)\Bigg]\Bigg\}.
+\end{align}
+%
+A C implementation using Horner's scheme could read
+%
+\begin{lstlisting}[language=C]
+double log1p_horner(double x) {
+   return x*(1 + x*(-0.5 + x*(1./3
+     + x*(-0.25 + x*(0.2
+     + x*(-1./6 + x/7))))));
+}
+\end{lstlisting}
+
+Several modern CPU architectures support instruction-level parallelism
+(ILP), which allows for the execution of multiple instructions at the
+same time.  To make use of ILP to reduce the execution time of the
+function, one could use Estrin's scheme \cite{estrin} to rewrite the
+function as
+%
+\begin{align}
+  \log(1+x) &\approx x + \left(\frac{x}{3}-\frac{1}{2}\right)
+    y+\left(\left(\frac{x}{7}-\frac{1}{6}\right)
+    y+\frac{x}{5}-\frac{1}{4}\right) z,
+\end{align}
+%
+where $y=x^2$ and $z=y^2$.  This form requires 8 floating-point
+multiplications and 6 additions, i.e.\ more multiplications than
+Horner's scheme.  However, in Estrin's form several terms can be
+executed in parallel, leading to a potential speed-up.  A C
+implementation using Estrin's scheme could read:
+%
+\begin{lstlisting}[language=C]
+double log1p_estrin(double x) {
+   const double y = x*x;
+   const double z = y*y;
+   return x + (x/3 - 0.5)*y
+      + (y*(-1./6 + x/7) + x/5 - 0.25)*z;
+}
+\end{lstlisting}
+
 \section{Conclusions}

### 2021-12-06 paper.tex
present Pade and mixed implementations

@@
   showstringspaces=false,
-  frame=shadowbox,
+  frame=single,
   rulesep=0.1em,
@@
   \log(1+x) \approx x - \frac{x^2}{2} + \frac{x^3}{3} - \frac{x^4}{4} + \frac{x^5}{5} - \frac{x^6}{6} + \frac{x^7}{7},
+  \label{eq:taylor}
 \end{align}
@@
   &+ x \Bigg[\frac{1}{5} + x\Bigg(-\frac{1}{6} + \frac{x}{7}\Bigg)\Bigg]\Bigg\}\Bigg)\Bigg]\Bigg\}.
+  \label{eq:horner}
 \end{align}
@@
     y+\frac{x}{5}-\frac{1}{4}\right) z,
+  \label{eq:estrin}
 \end{align}
@@
 
+To exploit ILP further, one could use a rational function
+approximation, such as a Padé approximant \cite{pade}, where the
+numerator and the denominator can be calculated in parallel.  A Padé
+approximant of $\log(1+x)$ at the same order as Eq.~\eqref{eq:taylor}
+reads
+%
+\begin{align}
+  \log(1+x) \approx x
+  \frac{\frac{40 x^3}{718121}+\frac{79040 x^2}{2154363}+\frac{517208
+    x}{718121}+1}{\frac{640 x^3}{718121}+\frac{225120 x^2}{718121}+\frac{876192
+    x}{718121}+1}.
+  \label{eq:pade}
+\end{align}
+%
+A C implementation of Eq.~\eqref{eq:pade} could read:
+%
+\begin{lstlisting}[language=C]
+double log1p_pade(double x) {
+   const double num = 1 + 517208./718121*x
+      + 79040./2154363*x*x + 40./718121*x*x*x;
+   const double den = 1 + 876192./718121*x
+      + 225120./718121*x*x + 640./718121*x*x*x;
+   return x*num/den;
+}
+\end{lstlisting}
+%
+The numerator and the denominator of Eq.~\eqref{eq:pade} can be
+re-written using Estrin's scheme, allowing for further ILP:
+%
+\begin{lstlisting}[language=C]
+double log1p_mixed(double x) {
+   const double y = x*x;
+   const double num = 1 + 517208./718121*x
+      + y*(79040./2154363 + 40./718121*x);
+   const double den = 1 + 876192./718121*x
+      + y*(225120./718121 + 640./718121*x);
+   return x*num/den;
+}
+\end{lstlisting}
+%
+Note that the value $y=x^2$ can be re-used in the calculation of both
+the numerator and in the denominator.  This ``mixed'' implementation
+requires 8 floating-point multiplications, 1 floating-point division
+and 6 additions.
+
+\section{Performance comparison}
+
 \section{Conclusions}

### 2021-12-06 paper.tex
improve affiliation
refine text

@@
 \author{\myauthor}
-\affil{Fachbereich Energie und Biotechnologie, Hochschule Flensburg, Kanzleistraße 91--93, 24943 Flensburg, Germany}
+\affil{Fachbereich Energie und Biotechnologie, Hochschule Flensburg,\\ Kanzleistra{\ss}e 91--93, 24943 Flensburg, Germany}
 \date{\today}
@@
 
-\section{Comparison of algorithms}
+\section{Performance comparison of different implementation strategies}
 
-As an example we consider the function $\log(1+x)$, which has the
-following Taylor expansion around $x=0$ for $-1 < x \leq 1$:
+As a toy example we consider the function $\log(1+x)$, which has the
+Taylor expansion
 %
 \begin{align}
-  \log(1+x) \approx x - \frac{x^2}{2} + \frac{x^3}{3} - \frac{x^4}{4} + \frac{x^5}{5} - \frac{x^6}{6} + \frac{x^7}{7},
+  \log(1+x) \approx x - \frac{x^2}{2} + \frac{x^3}{3} - \frac{x^4}{4} + \frac{x^5}{5} - \frac{x^6}{6} + \frac{x^7}{7}
   \label{eq:taylor}
@@
 %
-where terms of $O(x^8)$ have been omitted.
-In C this function could be implemented as follows:
+around $x=0$ for $x\in(-1,1]$, where terms of $O(x^8)$ and beyond have
+been omitted.  In the following we will briefly study various possible
+implementations of this function and compare the performance of the
+different implementations.  For brevity we will restrict ourselves to
+$x\in(-1,1]$.  In C the function in Eq.~\eqref{eq:taylor} could be
+implemented very naively as follows:
 %
@@
   $x/2$, as less expensive floating-point multiplication by a
-  constant, for example as $0.5x$.}
+  constant, for example as $0.5x$.  In the following we will therefore
+  count a division of a variable by a constant as a floating-point
+  multiplication.}
 

### 2021-12-06 paper.tex
adding word

@@
 
-As is well known, the performante of this implementation can be
+As is well known, the performante of this naive implementation can be
 improved by applying Horner's scheme, which reduces the number of

### 2021-12-06 paper.tex
break equation
use split
shorten implementation
improve section title
move source file to src/ directory

@@
 
-\section{Performance comparison of different implementation strategies}
+\section{Comparison of different implementation strategies}
 
@@
 %
-\begin{lstlisting}[language=C]
-double log1p_naive(double x) {
-   return x - x*x/2 + x*x*x/3 - x*x*x*x/4
-      + x*x*x*x*x/5 - x*x*x*x*x*x/6
-      + x*x*x*x*x*x*x/7;
-}
-\end{lstlisting}
+\lstinputlisting[language=C]{src/log1p_naive.c}
 %
@@
 \begin{align}
-  \log(1+x) \approx{}& x \Bigg\{1 + x \Bigg[-\frac{1}{2} + x \Bigg(\frac{1}{3} + x \Bigg\{-\frac{1}{4}  \notag \\
+\begin{split}
+  \log(1+x) \approx{}& x \Bigg\{1 + x \Bigg[-\frac{1}{2} + x \Bigg(\frac{1}{3} + x \Bigg\{-\frac{1}{4} \\
   &+ x \Bigg[\frac{1}{5} + x\Bigg(-\frac{1}{6} + \frac{x}{7}\Bigg)\Bigg]\Bigg\}\Bigg)\Bigg]\Bigg\}.
-  \label{eq:horner}
+\end{split}\label{eq:horner}%
 \end{align}
@@
 %
-\begin{lstlisting}[language=C]
-double log1p_horner(double x) {
-   return x*(1 + x*(-0.5 + x*(1./3
-     + x*(-0.25 + x*(0.2
-     + x*(-1./6 + x/7))))));
-}
-\end{lstlisting}
+\lstinputlisting[language=C]{src/log1p_horner.c}
 
@@
 \begin{align}
-  \log(1+x) &\approx x + \left(\frac{x}{3}-\frac{1}{2}\right)
-    y+\left(\left(\frac{x}{7}-\frac{1}{6}\right)
-    y+\frac{x}{5}-\frac{1}{4}\right) z,
-  \label{eq:estrin}
+\begin{split}
+  \log(1+x) \approx{}& x + \left(\frac{x}{3}-\frac{1}{2}\right)
+    y \\ &+\left[\left(\frac{x}{7}-\frac{1}{6}\right)
+    y+\frac{x}{5}-\frac{1}{4}\right] z,
+\end{split}\label{eq:estrin}%
 \end{align}
@@
 %
-\begin{lstlisting}[language=C]
-double log1p_estrin(double x) {
-   const double y = x*x;
-   const double z = y*y;
-   return x + (x/3 - 0.5)*y
-      + (y*(-1./6 + x/7) + x/5 - 0.25)*z;
-}
-\end{lstlisting}
+\lstinputlisting[language=C]{src/log1p_estrin.c}
 
@@
 %
-A C implementation of Eq.~\eqref{eq:pade} could read:
+A naive C implementation of Eq.~\eqref{eq:pade} could read:
 %
-\begin{lstlisting}[language=C]
-double log1p_pade(double x) {
-   const double num = 1 + 517208./718121*x
-      + 79040./2154363*x*x + 40./718121*x*x*x;
-   const double den = 1 + 876192./718121*x
-      + 225120./718121*x*x + 640./718121*x*x*x;
-   return x*num/den;
-}
-\end{lstlisting}
+\lstinputlisting[language=C]{src/log1p_pade.c}
 %
@@
 %
-\begin{lstlisting}[language=C]
-double log1p_mixed(double x) {
-   const double y = x*x;
-   const double num = 1 + 517208./718121*x
-      + y*(79040./2154363 + 40./718121*x);
-   const double den = 1 + 876192./718121*x
-      + y*(225120./718121 + 640./718121*x);
-   return x*num/den;
-}
-\end{lstlisting}
+\lstinputlisting[language=C]{src/log1p_mixed.c}
 %
@@
 
-\section{Performance comparison}
+\section{Application to the implementation of the real dilogarithm}
 

### 2021-12-06 src/log1p_estrin.c

@@ -1,0 +1,6 @@
+double log1p_estrin(double x) {
+   const double y = x*x;
+   const double z = y*y;
+   return x + (x/3 - 0.5)*y
+      + (y*(-1./6 + x/7) + x/5 - 0.25)*z;
+}

### 2021-12-06 src/log1p_horner.c

@@ -1,0 +1,5 @@
+double log1p_horner(double x) {
+   return x*(1 + x*(-0.5 + x*(1./3
+     + x*(-0.25 + x*(0.2
+     + x*(-1./6 + x/7))))));
+}

### 2021-12-06 src/log1p_mixed.c

@@ -1,0 +1,8 @@
+double log1p_mixed(double x) {
+   const double y = x*x;
+   const double num = 1 + 517208./718121*x +
+      y*(79040./2154363 + 40./718121*x);
+   const double den = 1 + 876192./718121*x +
+      y*(225120./718121 + 640./718121*x);
+   return x*num/den;
+}

### 2021-12-06 src/log1p_naive.c

@@ -1,0 +1,5 @@
+double log1p_naive(double x) {
+   return x - x*x/2 + x*x*x/3 - x*x*x*x/4
+      + x*x*x*x*x/5 - x*x*x*x*x*x/6
+      + x*x*x*x*x*x*x/7;
+}

### 2021-12-06 src/log1p_pade.c

@@ -1,0 +1,7 @@
+double log1p_pade(double x) {
+   const double num = 1 + 517208./718121*x +
+      79040./2154363*x*x + 40./718121*x*x*x;
+   const double den = 1 + 876192./718121*x +
+      225120./718121*x*x + 640./718121*x*x*x;
+   return x*num/den;
+}

### 2021-12-06 src/test_log1p.c
adding test for log1p implementations

@@ -1,0 +1,60 @@
+#include <stdio.h>
+#include <math.h>
+
+unsigned failed = 0;
+unsigned passed = 0;
+
+double log1p_naive(double);
+double log1p_horner(double);
+double log1p_estrin(double);
+double log1p_pade(double);
+double log1p_mixed(double);
+
+struct Fn {
+   double (*f)(double);
+   const char* name;
+};
+
+const struct Fn functions[] = {
+   { log1p_naive , "log1p_naive"  },
+   { log1p_horner, "log1p_horner" },
+   { log1p_estrin, "log1p_estrin" },
+   { log1p_pade  , "log1p_pade"   },
+   { log1p_mixed , "log1p_mixed"  },
+};
+
+
+const double values[] = {
+   0, 0.1
+};
+
+
+#define CHECK_CLOSE(val,a,b,eps,name) do {                              \
+      const double aa = fabs(a);                                        \
+      const double ab = fabs(b);                                        \
+      const double mx = aa > ab ? aa : ab;                              \
+      const double ae = fabs(eps);                                      \
+      if (fabs(aa - ab) < ae*(1 + mx)) {                                \
+         passed++;                                                      \
+      } else {                                                          \
+         printf("Error in %s(%g): %g ~!~ %g with eps = %g\n", name, val, aa, ab, ae); \
+         failed++;                                                      \
+      }                                                                 \
+   } while (0);
+
+
+int main() {
+   const double eps = 1e-6;
+
+   for (int fi = 0; fi < sizeof(functions)/sizeof(functions[0]); ++fi) {
+      for (int v = 0; v < sizeof(values)/sizeof(values[0]); ++v) {
+         const double (*f)(double) = functions[fi].f;
+         const double val = values[v];
+         CHECK_CLOSE(val, f(val), log1p(val), eps, functions[fi].name);
+      }
+   }
+
+   printf("Tests passed: %u/%u\n", passed, passed + failed);
+
+   return failed > 0;
+}

### 2021-12-06 src/test_log1p.c
rename variable

@@ -46,11 +46,11 @@ const double values[] = {
 int main() {
    const double eps = 1e-6;
 
-   for (int fi = 0; fi < sizeof(functions)/sizeof(functions[0]); ++fi) {
+   for (int i = 0; i < sizeof(functions)/sizeof(functions[0]); ++i) {
       for (int v = 0; v < sizeof(values)/sizeof(values[0]); ++v) {
-         const double (*f)(double) = functions[fi].f;
+         const double (*f)(double) = functions[i].f;
          const double val = values[v];
-         CHECK_CLOSE(val, f(val), log1p(val), eps, functions[fi].name);
+         CHECK_CLOSE(val, f(val), log1p(val), eps, functions[i].name);
       }
    }
 

### 2021-12-06 src/test_log1p.c
adding test value

@@ -25,7 +25,7 @@ const struct Fn functions[] = {
 
 
 const double values[] = {
-   0, 0.1
+   0, 0.1, -0.1
 };
 
 
@@ -44,7 +44,7 @@ const double values[] = {
 
 
 int main() {
-   const double eps = 1e-6;
+   const double eps = 1e-5;
 
    for (int i = 0; i < sizeof(functions)/sizeof(functions[0]); ++i) {
       for (int v = 0; v < sizeof(values)/sizeof(values[0]); ++v) {

### 2021-12-07 src/log1p_mixed.c
correcting Pade approximant

@@ -1,8 +1,8 @@
 double log1p_mixed(double x) {
    const double y = x*x;
-   const double num = 1 + 517208./718121*x +
-      y*(79040./2154363 + 40./718121*x);
-   const double den = 1 + 876192./718121*x +
-      y*(225120./718121 + 640./718121*x);
+   const double num = 1 + 51064./40143*x +
+      y*(44320./120429 + 320./40143*x);
+   const double den = 1 + 23712./13381*x +
+      y*(12320./13381 + 5120./40143*x);
    return x*num/den;
 }

### 2021-12-07 src/log1p_pade.c

@@ -1,7 +1,7 @@
 double log1p_pade(double x) {
-   const double num = 1 + 517208./718121*x +
-      79040./2154363*x*x + 40./718121*x*x*x;
-   const double den = 1 + 876192./718121*x +
-      225120./718121*x*x + 640./718121*x*x*x;
+   const double num = 1 + 51064./40143*x +
+      44320./120429*x*x + 320./40143*x*x*x;
+   const double den = 1 + 23712./13381*x +
+      12320./13381*x*x + 5120./40143*x*x*x;
    return x*num/den;
 }

### 2021-12-07 src/test_log1p.c

@@ -44,7 +44,7 @@ const double values[] = {
 
 
 int main() {
-   const double eps = 1e-5;
+   const double eps = 1e-6;
 
    for (int i = 0; i < sizeof(functions)/sizeof(functions[0]); ++i) {
       for (int v = 0; v < sizeof(values)/sizeof(values[0]); ++v) {

### 2021-12-07 src/test_log1p.c
more precise error message

@@ -34,10 +34,11 @@ const double values[] = {
       const double ab = fabs(b);                                        \
       const double mx = aa > ab ? aa : ab;                              \
       const double ae = fabs(eps);                                      \
-      if (fabs(aa - ab) < ae*(1 + mx)) {                                \
+      const double rel = fabs(aa - ab)/(1 + mx);                        \
+      if (rel < ae) {                                                   \
          passed++;                                                      \
       } else {                                                          \
-         printf("Error in %s(%g): %g ~!~ %g with eps = %g\n", name, val, aa, ab, ae); \
+         printf("Error in %s(%g): %g ~!~ %g (rel. diff. %g < %g)\n", name, val, aa, ab, rel, ae); \
          failed++;                                                      \
       }                                                                 \
    } while (0);

### 2021-12-07 paper.tex
improve title
correcting Pade approximant
allow hyphenation of words that contain a dash

@@
 \usepackage{listings}
+\usepackage[shortcuts]{extdash} % \-/
 \usepackage{hyperref}
 
-\newcommand{\mytitle}{Notes on the high-performance implementation of special functions}
+\newcommand{\mytitle}{The performance gain of instruction-level parallelism on the evaluation of special functions}
 \newcommand{\myauthor}{Alexander Voigt}
@@
 %
-This naive implementation requires 26 floating-point multiplications
-and 6 floating-point additions.\footnote{Modern compilers usually
-  rewrite expensive floating-point divisions by a constant, such as
-  $x/2$, as less expensive floating-point multiplication by a
+This naive implementation requires 26 floating\-/point multiplications
+and 6 floating\-/point additions.\footnote{Modern compilers usually
+  rewrite expensive floating\-/point divisions by a constant, such as
+  $x/2$, as less expensive floating\-/point multiplication by a
   constant, for example as $0.5x$.  In the following we will therefore
-  count a division of a variable by a constant as a floating-point
+  count a division of a variable by a constant as a floating\-/point
   multiplication.}
@@
 improved by applying Horner's scheme, which reduces the number of
-expensive floating-point multiplications to 7 and requires 6
+expensive floating\-/point multiplications to 7 and requires 6
 additions:
@@
 
-Several modern CPU architectures support instruction-level parallelism
+Several modern CPU architectures support instruction\-/level parallelism
 (ILP), which allows for the execution of multiple instructions at the
@@
 %
-where $y=x^2$ and $z=y^2$.  This form requires 8 floating-point
+where $y=x^2$ and $z=y^2$.  This form requires 8 floating\-/point
 multiplications and 6 additions, i.e.\ more multiplications than
@@
   \log(1+x) \approx x
-  \frac{\frac{40 x^3}{718121}+\frac{79040 x^2}{2154363}+\frac{517208
-    x}{718121}+1}{\frac{640 x^3}{718121}+\frac{225120 x^2}{718121}+\frac{876192
-    x}{718121}+1}.
+  \frac{\frac{320 x^3}{40143}+\frac{44320 x^2}{120429}+\frac{51064
+    x}{40143}+1}{\frac{5120 x^3}{40143}+\frac{12320 x^2}{13381}+\frac{23712
+    x}{13381}+1}.
   \label{eq:pade}
@@
 the numerator and in the denominator.  This ``mixed'' implementation
-requires 8 floating-point multiplications, 1 floating-point division
+requires 8 floating\-/point multiplications, 1 floating\-/point division
 and 6 additions.

### 2021-12-07 src/log1p.h
move function declarations into a header

@@ -1,0 +1,18 @@
+double log1p_naive(double);
+double log1p_horner(double);
+double log1p_estrin(double);
+double log1p_pade(double);
+double log1p_mixed(double);
+
+struct Fn {
+   double (*f)(double);
+   const char* name;
+};
+
+const struct Fn functions[] = {
+   { log1p_naive , "log1p_naive"  },
+   { log1p_horner, "log1p_horner" },
+   { log1p_estrin, "log1p_estrin" },
+   { log1p_pade  , "log1p_pade"   },
+   { log1p_mixed , "log1p_mixed"  },
+};

### 2021-12-07 src/test_log1p.c

@@ -1,27 +1,9 @@
 #include <stdio.h>
 #include <math.h>
+#include "log1p.h"
 
 unsigned failed = 0;
 unsigned passed = 0;
-
-double log1p_naive(double);
-double log1p_horner(double);
-double log1p_estrin(double);
-double log1p_pade(double);
-double log1p_mixed(double);
-
-struct Fn {
-   double (*f)(double);
-   const char* name;
-};
-
-const struct Fn functions[] = {
-   { log1p_naive , "log1p_naive"  },
-   { log1p_horner, "log1p_horner" },
-   { log1p_estrin, "log1p_estrin" },
-   { log1p_pade  , "log1p_pade"   },
-   { log1p_mixed , "log1p_mixed"  },
-};
 
 
 const double values[] = {

### 2021-12-07 src/pade.m
adding script to calculate Pade approximant

@@ -1,0 +1,12 @@
+Needs["FunctionApproximations`"];
+
+(* bring rational function to standard form *)
+PolynomialStandardForm[expr_, x_] :=
+    Module[{n = Numerator[expr], d = Denominator[expr], c},
+           c = d /. x -> 0;
+           Expand[n/c] / Expand[d/c]
+    ];
+
+pade = EconomizedRationalApproximation[Log[1+x]/x, {x, {-1,1}, 3, 3}]
+
+pade = PolynomialStandardForm[pade,x]

### 2021-12-07 src/bench_log1p.c
adding benchmark target

@@ -1,0 +1,35 @@
+#include <stdio.h>
+#include <stdlib.h>
+#include <math.h>
+#include <time.h>
+#include "log1p.h"
+
+#define N 1000000
+
+double values[N];
+
+
+double bench_fn(double (*f)(double), const char* name) {
+   const double time_in_s = 0;
+
+   for (int i = 0; i < sizeof(values)/sizeof(values[0]); ++i) {
+      double res = f(values[i]);
+   }
+
+   printf("%s: %g seconds\n", name, time_in_s);
+}
+
+
+int main() {
+   srand(time(NULL));
+
+   for (unsigned i = 0; i < N; ++i) {
+      values[i] = rand()/RAND_MAX;
+   }
+
+   for (int i = 0; i < sizeof(functions)/sizeof(functions[0]); ++i) {
+      bench_fn(functions[i].f, functions[i].name);
+   }
+
+   return 0;
+}

### 2021-12-07 src/bench_log1p.c
convert to fp value

@@ -24,7 +24,7 @@ int main() {
    srand(time(NULL));
 
    for (unsigned i = 0; i < N; ++i) {
-      values[i] = rand()/RAND_MAX;
+      values[i] = 1.0*rand()/RAND_MAX;
    }
 
    for (int i = 0; i < sizeof(functions)/sizeof(functions[0]); ++i) {

### 2021-12-07 paper.tex
extend text

@@
 \newcommand{\myauthor}{Alexander Voigt}
+\newcommand{\Li}{\operatorname{Li}_2}
 
@@
 
+As an application we consider the implementation of the real
+dilogarithm, which appears in many physics applications, for example
+in the calculation of quantum corrections in quantum electrodynamics.
+The real dilogarithm is defined as
+%
+\begin{align}
+  \Li(x) &= -\int_0^x \frac{\log(1-t)}{t}dt, &
+  &(x<1).
+\end{align}
+%
+For $|x|<1$ the real dilogarithm has the series representation
+%
+\begin{align}
+  \Li(x) = \sum_{k=1}^\infty \frac{x^k}{k^2}.
+\end{align}
+%
+Further series representations exists, for example
+%
+\begin{align}
+  \Li(x) = \sum_{k=0}^\infty \frac{B_k}{(k+1)!} [-\log(1-x)]^{k+1},
+\end{align}
+%
+where $B_k$ are the Bernoulli numbers.  A representation in terms
+of Chebyshev polynomials also exists \cite{luke},
+%
+\begin{align}
+  \Li(x) = \sum_{k=0}^\infty a_n T_n^*(x),
+\end{align}
+%
+which is used for example in the ROOT program library \cite{root}.
+
 \section{Conclusions}

### 2021-12-07 paper.tex
adding ref
refine text
extend text

@@
 
-Several modern CPU architectures support instruction\-/level parallelism
-(ILP), which allows for the execution of multiple instructions at the
-same time.  To make use of ILP to reduce the execution time of the
-function, one could use Estrin's scheme \cite{estrin} to rewrite the
-function as
+Several modern CPU architectures support instruction\-/level
+parallelism (ILP), which allows for the execution of multiple
+instructions at the same time.  To make use of ILP, one could use
+Estrin's scheme \cite{estrin} to rewrite the function as
 %
@@
 Horner's scheme.  However, in Estrin's form several terms can be
-executed in parallel, leading to a potential speed-up.  A C
-implementation using Estrin's scheme could read:
+executed in parallel, leading to a potential speed-up if the cost of
+the additional multiplications is smaller than the gain by the
+parallel execution.  A C implementation using Estrin's scheme could
+read:
 %
@@
 %
+\begin{align}
+  \log(1+x) \approx x
+  \frac{\left(\frac{320 x}{40143}+\frac{44320}{120429}\right)y+\frac{51064
+    x}{40143}+1}{\left(\frac{5120 x}{40143}+\frac{12320}{13381}\right)y+\frac{23712
+    x}{13381}+1},
+  \label{eq:mixed}
+\end{align}
+%
+where $y=x^2$.  A C implementation could read:
+%
 \lstinputlisting[language=C]{src/log1p_mixed.c}
@@
 the numerator and in the denominator.  This ``mixed'' implementation
-requires 8 floating\-/point multiplications, 1 floating\-/point division
-and 6 additions.
+requires 8 floating\-/point multiplications, 1 floating\-/point
+division and 6 additions.  The rational function approximation
+\eqref{eq:mixed} may be faster for long numerator/denominator
+polynomials, where the cost of the floating\-/point division is
+smaller than the gain by the parallel execution.
 
@@
 %
-Further series representations exists, for example
+Further series representations exist, for example the accelerated
+series
 %
@@
 %
-where $B_k$ are the Bernoulli numbers.  A representation in terms
-of Chebyshev polynomials also exists \cite{luke},
+where $B_k$ are the Bernoulli numbers.  A series representation in
+terms of Chebyshev polynomials of the first kind $T_n(x)$ is given in
+\cite{luke},
 %
 \begin{align}
-  \Li(x) = \sum_{k=0}^\infty a_n T_n^*(x),
+  \Li(x) = \sum_{k=0}^\infty a_n T_n(2x-1),
+  \label{eq:cheby}
 \end{align}

### 2021-12-07 paper.tex
extend text

@@
   \Li(x) = \sum_{k=1}^\infty \frac{x^k}{k^2}.
+  \label{eq:series}
 \end{align}
@@
 which is used for example in the ROOT program library \cite{root}.
+Eqs.~\eqref{eq:series}--\eqref{eq:cheby} can be summed, for example
+using Horner's scheme or Clenshaw's algorithm \cite{clenshaw}.  These
+summation techniques are, however, purely sequential and thus do not
+make use of ILP.  In the following an approximation of the real
+dilogarithm is presented, which makes use of ILP, using a rational
+function approximation in combination with Estrin's scheme, similarly
+to the approach use in Eq.~\eqref{eq:mixed}.
 

### 2021-12-07 paper.tex
adding source code for Li2

@@
 \usepackage[shortcuts]{extdash} % \-/
+\usepackage{xcolor}
 \usepackage{hyperref}
@@
   stringstyle=\color{red},
-  commentstyle=\color{green},
+  commentstyle=\color{gray},
 }
@@
 
+\appendix
+
+\section{Implementation of the dilogarithm}
+
+\lstinputlisting[language=C]{src/Li2.c}
+
 \bibliographystyle{ieeetr}

### 2021-12-07 src/Li2.c

@@ -1,0 +1,79 @@
+#include <math.h>
+
+double li2(double x)
+{
+   const double PI = 3.1415926535897932;
+   const double P[] = {
+      1.0706105563309304277e+0,
+     -4.5353562730201404017e+0,
+      7.4819657596286408905e+0,
+     -6.0516124315132409155e+0,
+      2.4733515209909815443e+0,
+     -4.6937565143754629578e-1,
+      3.1608910440687221695e-2,
+     -2.4630612614645039828e-4
+   };
+   const double Q[] = {
+      1.0000000000000000000e+0,
+     -4.5355682121856044935e+0,
+      8.1790029773247428573e+0,
+     -7.4634190853767468810e+0,
+      3.6245392503925290187e+0,
+     -8.9936784740041174897e-1,
+      9.8554565816757007266e-2,
+     -3.2116618742475189569e-3
+   };
+
+   double y = 0, r = 0, s = 1;
+
+   /* transform to [0, 1/2) */
+   if (x < -1) {
+      const double l = log(1 - x);
+      y = 1/(1 - x);
+      r = -PI*PI/6 + l*(0.5*l - log(-x));
+      s = 1;
+   } else if (x == -1) {
+      return -PI*PI/12;
+   } else if (x < 0) {
+      const double l = log1p(-x);
+      y = x/(x - 1);
+      r = -0.5*l*l;
+      s = -1;
+   } else if (x == 0) {
+      return 0;
+   } else if (x < 0.5) {
+      y = x;
+      r = 0;
+      s = 1;
+   } else if (x < 1) {
+      y = 1 - x;
+      r = PI*PI/6 - log(x)*log(1 - x);
+      s = -1;
+   } else if (x == 1) {
+      return PI*PI/6;
+   } else if (x < 2) {
+      const double l = log(x);
+      y = 1 - 1/x;
+      r = PI*PI/6 - l*(log(1 - 1/x) + 0.5*l);
+      s = 1;
+   } else {
+      const double l = log(x);
+      y = 1/x;
+      r = PI*PI/3 - 0.5*l*l;
+      s = -1;
+   }
+
+   const double z = y - 0.25;
+   const double z2 = z*z;
+   const double z4 = z2*z2;
+   const double p = P[0] + z * P[1]
+      + z2 * (P[2] + z * P[3])
+      + z4 * (P[4] + z * P[5]
[…]

### 2021-12-07 paper.tex
extend section title
extend text

@@
 \usepackage{lmodern}
-\usepackage{amsmath}
+\usepackage{amsmath,amssymb}
 \usepackage[german]{authblk}
@@
 which is used for example in the ROOT program library \cite{root}.
-Eqs.~\eqref{eq:series}--\eqref{eq:cheby} can be summed, for example
-using Horner's scheme or Clenshaw's algorithm \cite{clenshaw}.  These
-summation techniques are, however, purely sequential and thus do not
-make use of ILP.  In the following an approximation of the real
-dilogarithm is presented, which makes use of ILP, using a rational
-function approximation in combination with Estrin's scheme, similarly
-to the approach use in Eq.~\eqref{eq:mixed}.
+The r.h.s.\ of Eqs.~\eqref{eq:series}--\eqref{eq:cheby} can be summed,
+for example, using Horner's scheme or Clenshaw's algorithm
+\cite{clenshaw}.  These summation techniques are, however, purely
+sequential and thus do not make use of ILP.  In the following an
+approximation of the real dilogarithm is presented, which makes use of
+ILP, using a rational function approximation in combination with
+Estrin's scheme, similarly to the approach use in
+Eq.~\eqref{eq:mixed}.
+
+We will define the real dilogarithm $\Li(x)$ for all $x\in\mathbb{R}$
+as the real part of the complex continuation $\Li(z)$,
+$z\in\mathbb{C}$.  The function can be mapped onto the interval
+$[0,1/2)$ by using the following identities:
 
@@
 
-\section{Implementation of the dilogarithm}
+\section{Implementation of the real dilogarithm}
 

### 2021-12-07 paper.tex
adding identities

@@
 dilogarithm, which appears in many physics applications, for example
-in the calculation of quantum corrections in quantum electrodynamics.
+in the calculation of quantum corrections in quantum field theory.
 The real dilogarithm is defined as
@@
 We will define the real dilogarithm $\Li(x)$ for all $x\in\mathbb{R}$
-as the real part of the complex continuation $\Li(z)$,
-$z\in\mathbb{C}$.  The function can be mapped onto the interval
-$[0,1/2)$ by using the following identities:
+as the real part of the complex continuation of the dilogarithm.  The
+real dilogarithm can be mapped onto the interval $[0,1/2)$ by using
+the following identities:
+%
+\begin{align}
+  \Li(x) &= -\Li\left(\frac{1}{1-x}\right) - \frac{\pi^2}{6} + \log(1-x)\left[\frac{1}{2}\log(1-x) - \log(-x)\right], \quad (x<-1), \\
+  \Li(x) &= -\Li\left(\frac{x}{x-1}\right) - \frac{1}{2}\log^2(x), \quad (-1<x<0), \\
+  \Li(x) &= -\Li(1-x) + \frac{\pi^2}{6} - \log(x) \log(1-x), \quad (\frac{1}{2}\leq x<1), \\
+  \Li(x) &= \Li\left(1-\frac{1}{x}\right) + \frac{\pi^2}{6} - \log(x)\left[\log\left(1-\frac{1}{x}\right) + \frac{1}{2}\log(x)\right], \quad (1<x<2), \\
+  \Li(x) &= -\Li\left(\frac{1}{x}\right) + \frac{\pi^2}{3} - \frac{1}{2} \log^2(x), \quad (x\geq 2).
+\end{align}
 

### 2021-12-07 paper.tex
use widetext for longish equations
improve text

@@
 \usepackage{xcolor}
+\usepackage{widetext}
 \usepackage{hyperref}
@@
 We will define the real dilogarithm $\Li(x)$ for all $x\in\mathbb{R}$
-as the real part of the complex continuation of the dilogarithm.  The
-real dilogarithm can be mapped onto the interval $[0,1/2)$ by using
-the following identities:
+as the real part of its complex continuation.  The real dilogarithm
+can be mapped onto the interval $[0,1/2]$ by using the following
+identities:
 %
+\begin{widetext}
 \begin{align}
@@
 \end{align}
+\end{widetext}
 

### 2021-12-08 paper.tex
alignment of ranges
extend text
fix typos
log -> ln

@@
 
-As a toy example we consider the function $\log(1+x)$, which has the
+As a toy example we consider the function $\ln(1+x)$, which has the
 Taylor expansion
@@
 \begin{align}
-  \log(1+x) \approx x - \frac{x^2}{2} + \frac{x^3}{3} - \frac{x^4}{4} + \frac{x^5}{5} - \frac{x^6}{6} + \frac{x^7}{7}
+  \ln(1+x) \approx x - \frac{x^2}{2} + \frac{x^3}{3} - \frac{x^4}{4} + \frac{x^5}{5} - \frac{x^6}{6} + \frac{x^7}{7}
   \label{eq:taylor}
@@
 \begin{split}
-  \log(1+x) \approx{}& x \Bigg\{1 + x \Bigg[-\frac{1}{2} + x \Bigg(\frac{1}{3} + x \Bigg\{-\frac{1}{4} \\
+  \ln(1+x) \approx{}& x \Bigg\{1 + x \Bigg[-\frac{1}{2} + x \Bigg(\frac{1}{3} + x \Bigg\{-\frac{1}{4} \\
   &+ x \Bigg[\frac{1}{5} + x\Bigg(-\frac{1}{6} + \frac{x}{7}\Bigg)\Bigg]\Bigg\}\Bigg)\Bigg]\Bigg\}.
@@
 \begin{split}
-  \log(1+x) \approx{}& x + \left(\frac{x}{3}-\frac{1}{2}\right)
+  \ln(1+x) \approx{}& x + \left(\frac{x}{3}-\frac{1}{2}\right)
     y \\ &+\left[\left(\frac{x}{7}-\frac{1}{6}\right)
@@
 numerator and the denominator can be calculated in parallel.  A Padé
-approximant of $\log(1+x)$ at the same order as Eq.~\eqref{eq:taylor}
+approximant of $\ln(1+x)$ at the same order as Eq.~\eqref{eq:taylor}
 reads
@@
 \begin{align}
-  \log(1+x) \approx x
+  \ln(1+x) \approx x
   \frac{\frac{320 x^3}{40143}+\frac{44320 x^2}{120429}+\frac{51064
@@
 \begin{align}
-  \log(1+x) \approx x
+  \ln(1+x) \approx x
   \frac{\left(\frac{320 x}{40143}+\frac{44320}{120429}\right)y+\frac{51064
@@
 \begin{align}
-  \Li(x) &= -\int_0^x \frac{\log(1-t)}{t}dt, &
+  \Li(x) &= -\int_0^x \frac{\ln(1-t)}{t}dt, &
   &(x<1).
@@
 \begin{align}
-  \Li(x) = \sum_{k=0}^\infty \frac{B_k}{(k+1)!} [-\log(1-x)]^{k+1},
+  \Li(x) = \sum_{k=0}^\infty \frac{B_k}{(k+1)!} [-\ln(1-x)]^{k+1},
 \end{align}
@@
 %
-\begin{widetext}
-\begin{align}
-  \Li(x) &= -\Li\left(\frac{1}{1-x}\right) - \frac{\pi^2}{6} + \log(1-x)\left[\frac{1}{2}\log(1-x) - \log(-x)\right], \quad (x<-1), \\
-  \Li(x) &= -\Li\left(\frac{x}{x-1}\right) - \frac{1}{2}\log^2(x), \quad (-1<x<0), \\
-  \Li(x) &= -\Li(1-x) + \frac{\pi^2}{6} - \log(x) \log(1-x), \quad (\frac{1}{2}\leq x<1), \\
-  \Li(x) &= \Li\left(1-\frac{1}{x}\right) + \frac{\pi^2}{6} - \log(x)\left[\log\left(1-\frac{1}{x}\right) + \frac{1}{2}\log(x)\right], \quad (1<x<2), \\
-  \Li(x) &= -\Li\left(\frac{1}{x}\right) + \frac{\pi^2}{3} - \frac{1}{2} \log^2(x), \quad (x\geq 2).
-\end{align}
-\end{widetext}
+\begin{widetext}%
+\begin{align}%
+  \Li(x) &= \Li\left(\frac{1}{1-x}\right) - \frac{\pi^2}{6} + \ln(1-x)\left[\frac{1}{2}\ln(1-x) - \ln(-x)\right], && (x<-1), \\
+  \Li(x) &= -\Li\left(\frac{x}{x-1}\right) - \frac{1}{2}\ln^2(1-x), && (-1<x<0), \\
+  \Li(x) &= -\Li(1-x) + \frac{\pi^2}{6} - \ln(x) \ln(1-x), && (\frac{1}{2}\leq x<1), \\
+  \Li(x) &= \Li\left(1-\frac{1}{x}\right) + \frac{\pi^2}{6} - \ln(x)\left[\ln\left(1-\frac{1}{x}\right) + \frac{1}{2}\ln(x)\right], && (1<x<2), \\
+  \Li(x) &= -\Li\left(\frac{1}{x}\right) + \frac{\pi^2}{3} - \frac{1}{2} \ln^2(x), && (x\geq 2),%
+\end{align}%
+\end{widetext}%
+%
+as well as the special values $\Li(0)=0$, $\Li(1)=\pi^2/6$ and
+$\Li(-1)=-\pi^2/12$.
 

### 2021-12-08 paper.tex
adapt directory
extend text

@@
 
+An algorithm for the computation of the real dilogarithm is presented,
+which allows for the use of instruction\-/level parallelism.
+
 \section{Introduction}
 
-\cite{polylogarithm}
+The dilogarithm \cite{lewin} is a special function, which appears in
+many physics applications, for example in the calculation of quantum
+corrections in quantum field theory.  The investigation of new models
+beyond the Standard Model of particle physics with a large parameter
+space require the evaluation of the dilogarithm (and other functions)
+billions of times.  Such investigations thus require a fast evaluation
+of the dilogarithm.  In the past many time-efficient algorithms have
+been proposed \cite{}, which are implemented in program libraries for
+numerical physics and mathematics \cite{root,gsl}.
+
+The development of modern CPUs with support for instruction\-/level
+parallelism (ILP) opens the opportunity for a further reduction of the
+runtime of the calculation of the dilogarithm, given an appropriate
+formulation of the algorithm.  In this paper an algorithm for the
+calculation of the real dilogarithm $\Li$ is presented, which makes
+use of ILP, resulting in a reduced runtime compared to the algorithms
+presented so far.  A C implementation of this algorithm can be found
+in the appendix and in the arXiv submission of this paper.
 
@@
 
+C, C++ and FORTRAN implementations of the presented algorithm can be
+found in the Polylogarithm package \cite{polylogarithm}.
+
 \appendix
@@
 
-\lstinputlisting[language=C]{src/Li2.c}
+\lstinputlisting[language=C]{anc/Li2.c}
 

### 2021-12-08 paper.tex
refine title
adding citation

@@
 
-\newcommand{\mytitle}{The performance gain of instruction-level parallelism on the evaluation of special functions}
+\newcommand{\mytitle}{An algorithm for the calculation of the real dilogarithm with instruction-level parallelism}
 \newcommand{\myauthor}{Alexander Voigt}
@@
 of the dilogarithm.  In the past many time-efficient algorithms have
-been proposed \cite{}, which are implemented in program libraries for
-numerical physics and mathematics \cite{root,gsl}.
+been proposed \cite{}, which are implemented, for example, in program
+libraries for numerical physics and mathematics
+\cite{root,gsl,cephes}.
 

### 2021-12-08 paper.tex
adding references

@@
-\documentclass[10pt,DIV16,twocolumn]{scrartcl}
+\documentclass[10pt,DIV16,twocolumn,numbers=noenddot]{scrartcl}
 
@@
 of the dilogarithm.  In the past many time-efficient algorithms have
-been proposed \cite{}, which are implemented, for example, in program
-libraries for numerical physics and mathematics
-\cite{root,gsl,cephes}.
+been invented (see e.g.~\cite{koelbigDilog,ginsberg,morris}) and many
+are implemented, for example, in general-purpose physics and
+mathematics program libraries \cite{root,gsl,cephes}.
 
@@
 C, C++ and FORTRAN implementations of the presented algorithm can be
-found in the Polylogarithm package \cite{polylogarithm}.
+found in the Polylogarithm library \cite{polylogarithm}.
 

### 2021-12-08 paper.tex
reformulation

@@
 
-\section{Application to the implementation of the real dilogarithm}
+\section{An ILP-aware algorithm for the real dilogarithm}
 
-As an application we consider the implementation of the real
-dilogarithm, which appears in many physics applications, for example
-in the calculation of quantum corrections in quantum field theory.
-The real dilogarithm is defined as
+In the following an ILP-aware implementation of the real dilogarithm
+is presented, which allows for the parallel execution of different
+floating\-/point operations.  The real dilogarithm is defined as
 %
@@
 
-We will define the real dilogarithm $\Li(x)$ for all $x\in\mathbb{R}$
-as the real part of its complex continuation.  The real dilogarithm
-can be mapped onto the interval $[0,1/2]$ by using the following
+We define the real dilogarithm $\Li(x)$ for all $x\in\mathbb{R}$ as
+the real part of its complex continuation.  The real dilogarithm can
+be mapped onto the interval $[0,1/2]$ by using the following
 identities:

### 2021-12-09 anc/Li2.c
typo

@@ -26,7 +26,7 @@ double li2(double x)
 
    double y = 0, r = 0, s = 1;
 
-   /* transform to [0, 1/2) */
+   /* transform to [0, 1/2] */
    if (x < -1) {
       const double l = log(1 - x);
       y = 1/(1 - x);

### 2021-12-09 paper.tex
adding footnote

@@
 ILP, using a rational function approximation in combination with
-Estrin's scheme, similarly to the approach use in
-Eq.~\eqref{eq:mixed}.
+Estrin's scheme, similar to the approach used in
+Eq.~\eqref{eq:mixed}.%
+\footnote{The Cephes library \cite{cephes} implements an equivalent
+  rational function approximation for Spence's function
+  $\operatorname{Sp}(x)=\Li(1-x)$.}
 

### 2021-12-09 paper.tex
adding references
extend intro

@@
 parallelism (ILP) opens the opportunity for a further reduction of the
-runtime of the calculation of the dilogarithm, given an appropriate
-formulation of the algorithm.  In this paper an algorithm for the
+runtime of the calculation of the dilogarithm.  To make use of ILP,
+the algorithm to calculate the dilogarithm must be formulated
+appropriately in order to allow for the parallel execution of multiple
+floating\-/point operations.  In this paper an algorithm for the
 calculation of the real dilogarithm $\Li$ is presented, which makes
@@
 found in the Polylogarithm library \cite{polylogarithm}.
+Implementations in Julia and Rust can be found in \cite{PolyLog.jl}
+and \cite{polylog}, respectively.
 

### 2021-12-09 paper.tex
extend intro
reformulation

@@
 \newcommand{\Li}{\operatorname{Li}_2}
+\newcommand{\secref}[1]{Section~\ref{#1}}
 
@@
 use of ILP, resulting in a reduced runtime compared to the algorithms
-presented so far.  A C implementation of this algorithm can be found
-in the appendix and in the arXiv submission of this paper.
+presented so far.
+
+This paper is organized as follows: In \secref{sec:toymodel} different
+implementation strategies are discussed using a toy example function.
+In \secref{sec:algorithm} an ILP-aware algorithm for the numerical
+calculation of the dilogarithm is presented.  A C implementation of
+this algorithm can be found in the appendix and in the arXiv
+submission of this paper.
 
 \section{Comparison of different implementation strategies}
+\label{sec:toymodel}
 
@@
 around $x=0$ for $x\in(-1,1]$, where terms of $O(x^8)$ and beyond have
-been omitted.  In the following we will briefly study various possible
-implementations of this function and compare the performance of the
+been omitted.  In the following we will briefly show various possible
+implementations of this function and discuss the performance of the
 different implementations.  For brevity we will restrict ourselves to
@@
 \section{An ILP-aware algorithm for the real dilogarithm}
+\label{sec:algorithm}
 

### 2021-12-09 paper.tex
shorten section title
use ordinary CM fonts
adding runtime table

@@
 \usepackage{xcolor}
+\usepackage{booktabs}
 \usepackage{widetext}
 \usepackage{hyperref}
+\setkomafont{disposition}{\bfseries}
 
@@
 \newcommand{\secref}[1]{Section~\ref{#1}}
+\newcommand{\tabref}[1]{\tablename~\ref{#1}}
 
@@
 
-\section{Comparison of different implementation strategies}
+\section{Comparison of implementation strategies}
 \label{sec:toymodel}
@@
 
+In \tabref{tab:runtime} the runtime of different C implementations is
+compared on two different CPU architectures.  On each architecture the
+implentations are compiled with \texttt{g++} 10.2.1 with optimization
+\texttt{-O2}.  The shown runtime denotes the total wall clock time for
+the calculation of $\Li(x)$ for $10^6$ uniformly distributed random
+values $x\in[-5,5]$.
+%
+\begin{table}[tb]
+  \centering
+  \caption{Runtime in seconds for the calculation of $\Li(x)$ for
+    different implementations on different CPU architectures.}
+  \begin{tabular}{lc}
+    \toprule
+    Implementation & i5-8265U \\
+    \midrule
+    this paper & $0.022$ \\
+    Cephes \cite{cephes} & $0.028$ \\
+    Algorithm 327 \cite{koelbigDilog} & $0.031$ \\
+    Chebyshev \cite{luke} & $0.058$ \\
+    Algorithm 490 \cite{ginsberg} & $0.059$ \\
+    \bottomrule
+  \end{tabular}
+  \label{tab:runtime}
+\end{table}
+%
+We find that the implementation presented in this paper is fastest
+among the ones shown in the table.  The next-to fastest implementation
+is provided by Cephes, which is only $\sim 20\%$ slower.  The short
+runtime of Cephes stems from the fact that Cephes also uses a rational
+function approximation, where the numerator and the denominator can be
+evaluated in parallel.  However, Cephes evaluates the polynomials in
+the numerator and denominator using Horner's scheme, which is purely
+sequential.  Furthermore, for $x>1$ Cephes calculates $\ln(x)$ twice,
+which leads to a further slowdown.
+
 \section{Conclusions}

### 2021-12-09 paper.tex
adding line

@@
     this paper & $0.022$ \\
+    Morris \cite{morris} & $0.022$ \\
     Cephes \cite{cephes} & $0.028$ \\

### 2021-12-09 paper.tex
extend text

@@
 as well as the special values $\Li(0)=0$, $\Li(1)=\pi^2/6$ and
-$\Li(-1)=-\pi^2/12$.
+$\Li(-1)=-\pi^2/12$.  Within the interval $[0,1/2]$ the real
+dilogarithm has the series expansion \eqref{eq:series}, which can be
+written as
+%
+\begin{align}
+  \Li(x) = x \sum_{k=0} \frac{x^{k}}{(k+1)^2}.
+  \label{eq:series2}
+\end{align}
+%
+We approximate the sum on the r.h.s.\ of Eq.~\eqref{eq:series2} using
+an economized Padé approximant on the interval $[0,1/2]$ using the
+\texttt{EconomizedRationalApproximation} from Wolfram/Mathematica
+\cite{mathematica},
+%
+\begin{align}
+  \sum_{k=0} \frac{x^{k}}{(k+1)^2} =
+  \frac{\sum_{i=0}^7 a_iy^i}{\sum_{j=0}^7 b_jy^j},
+\end{align}
+%
+where $y=1/4$ is the midpoint of the interval $[0,1/2]$.  The
+coefficients $a_i$ and $b_j$ ($i,j=0,\ldots,7$) are listed in
+\tabref{tab:coeffs}.
 

### 2021-12-10 paper.tex
extend table

@@
     different implementations on different CPU architectures.}
-  \begin{tabular}{lc}
+  \begin{tabular}{lcc}
     \toprule
-    Implementation & i5-8265U \\
+    Implementation & i5-8265U & i7-5600U \\
     \midrule
-    this paper & $0.022$ \\
-    Morris \cite{morris} & $0.022$ \\
-    Cephes \cite{cephes} & $0.028$ \\
-    Algorithm 327 \cite{koelbigDilog} & $0.031$ \\
-    Chebyshev \cite{luke} & $0.058$ \\
-    Algorithm 490 \cite{ginsberg} & $0.059$ \\
+    this paper & $0.022$ & $0.027$ \\
+    Morris \cite{morris} & $0.022$ & $0.027$ \\
+    Cephes \cite{cephes} & $0.028$ & $0.034$ \\
+    Algorithm 327 \cite{koelbigDilog} & $0.031$ & $0.035$ \\
+    Chebyshev \cite{luke} & $0.058$ & $0.060$ \\
+    Algorithm 490 \cite{ginsberg} & $0.059$ & $0.071$ \\
+    GSL \cite{gsl} & & $0.155$ \\
     \bottomrule

### 2021-12-10 paper.tex
extend table description

@@
     this paper & $0.022$ & $0.027$ \\
-    Morris \cite{morris} & $0.022$ & $0.027$ \\
+    Morris (DILOG 0011) \cite{morris} & $0.022$ & $0.027$ \\
     Cephes \cite{cephes} & $0.028$ & $0.034$ \\

### 2021-12-10 paper.tex
adding table with coeffs
update data for Morris,
who uses fewer terms in the numerator and denominator

--- 2021-12-10 paper.tex
@@
   \sum_{k=0} \frac{x^{k}}{(k+1)^2} =
-  \frac{\sum_{i=0}^7 a_iy^i}{\sum_{j=0}^7 b_jy^j},
+  \frac{\sum_{k=0}^7 a_ky^k}{\sum_{k=0}^7 b_ky^k},
+  \label{eq:pade2}
 \end{align}
@@
 where $y=1/4$ is the midpoint of the interval $[0,1/2]$.  The
-coefficients $a_i$ and $b_j$ ($i,j=0,\ldots,7$) are listed in
+coefficients $a_k$ and $b_k$ ($k=0,\ldots,7$) are listed in
 \tabref{tab:coeffs}.
+%
+\begin{table}[t]
+  \centering
+  \caption{Coefficients of the numerator and denominator polynomials
+    for the Padé approximant \eqref{eq:pade2}.}
+  \begin{tabular}{lr}
+    \toprule
+    $a_0$ & $ 1.0706105563309304277\cdot 10^{+0}$ \\
+    $a_1$ & $-4.5353562730201404017\cdot 10^{+0}$ \\
+    $a_2$ & $ 7.4819657596286408905\cdot 10^{+0}$ \\
+    $a_3$ & $-6.0516124315132409155\cdot 10^{+0}$ \\
+    $a_4$ & $ 2.4733515209909815443\cdot 10^{+0}$ \\
+    $a_5$ & $-4.6937565143754629578\cdot 10^{-1}$ \\
+    $a_6$ & $ 3.1608910440687221695\cdot 10^{-2}$ \\
+    $a_7$ & $-2.4630612614645039828\cdot 10^{-4}$ \\
+    $b_0$ & $ 1.0000000000000000000\cdot 10^{+0}$ \\
+    $b_1$ & $-4.5355682121856044935\cdot 10^{+0}$ \\
+    $b_2$ & $ 8.1790029773247428573\cdot 10^{+0}$ \\
+    $b_3$ & $-7.4634190853767468810\cdot 10^{+0}$ \\
+    $b_4$ & $ 3.6245392503925290187\cdot 10^{+0}$ \\
+    $b_5$ & $-8.9936784740041174897\cdot 10^{-1}$ \\
+    $b_6$ & $ 9.8554565816757007266\cdot 10^{-2}$ \\
+    $b_7$ & $-3.2116618742475189569\cdot 10^{-3}$ \\
+    \bottomrule
+  \end{tabular}
+  \label{tab:coeffs}
+\end{table}
 
@@
     this paper & $0.022$ & $0.027$ \\
-    Morris \cite{morris} & $0.022$ & $0.027$ \\
+    Morris (DILOG 0011) \cite{morris} & $0.022$ & $0.026$ \\
     Cephes \cite{cephes} & $0.028$ & $0.034$ \\

### 2021-12-12 paper.tex
rename symbols to avoid confusion
reduce number of coefficients

--- 2021-12-10 paper.tex
@@
 We approximate the sum on the r.h.s.\ of Eq.~\eqref{eq:series2} using
-an economized Padé approximant on the interval $[0,1/2]$ using the
-\texttt{EconomizedRationalApproximation} from Wolfram/Mathematica
+a rational minimax approximant on the interval $[0,1/2]$ using the
+\texttt{MiniMaxApproximation} from Wolfram/Mathematica
 \cite{mathematica},
@@
   \sum_{k=0} \frac{x^{k}}{(k+1)^2} =
-  \frac{\sum_{i=0}^7 a_iy^i}{\sum_{j=0}^7 b_jy^j},
+  \frac{\sum_{k=0}^5 p_kx^k}{\sum_{k=0}^6 q_kx^k}.
+  \label{eq:minimax}
 \end{align}
 %
-where $y=1/4$ is the midpoint of the interval $[0,1/2]$.  The
-coefficients $a_i$ and $b_j$ ($i,j=0,\ldots,7$) are listed in
-\tabref{tab:coeffs}.
+The coefficients $p_k$ and $q_k$ are listed in \tabref{tab:coeffs}.
+%
+\begin{table}[t]
+  \centering
+  \caption{Coefficients of the numerator and denominator polynomials
+    for the minimax approximant \eqref{eq:minimax}.}
+  \begin{tabular}{lr}
+    \toprule
+    $p_0$ & $ 0.9999999999999999502\cdot 10^{+0}$ \\
+    $p_1$ & $-2.6883926818565423430\cdot 10^{+0}$ \\
+    $p_2$ & $ 2.6477222699473109692\cdot 10^{+0}$ \\
+    $p_3$ & $-1.1538559607887416355\cdot 10^{+0}$ \\
+    $p_4$ & $ 2.0886077795020607837\cdot 10^{-1}$ \\
+    $p_5$ & $-1.0859777134152463084\cdot 10^{-2}$ \\
+    $q_0$ & $ 1.0000000000000000000\cdot 10^{+0}$ \\
+    $q_1$ & $-2.9383926818565635485\cdot 10^{+0}$ \\
+    $q_2$ & $ 3.2712093293018635389\cdot 10^{+0}$ \\
+    $q_3$ & $-1.7076702173954289421\cdot 10^{+0}$ \\
+    $q_4$ & $ 4.1596017228400603836\cdot 10^{-1}$ \\
+    $q_5$ & $-3.9801343754084482956\cdot 10^{-2}$ \\
+    $q_6$ & $ 8.2743668974466659035\cdot 10^{-4}$ \\
+    \bottomrule
+  \end{tabular}
+  \label{tab:coeffs}
+\end{table}
 
@@
     \midrule
-    this paper & $0.022$ & $0.027$ \\
-    Morris \cite{morris} & $0.022$ & $0.027$ \\
+    this paper & $0.022$ & $0.026$ \\
+    Morris (DILOG 0011) \cite{morris} & $0.022$ & $0.026$ \\
     Cephes \cite{cephes} & $0.028$ & $0.034$ \\

### 2021-12-12 anc/Li2.c

@@ -4,24 +4,21 @@ double li2(double x)
 {
    const double PI = 3.1415926535897932;
    const double P[] = {
-      1.0706105563309304277e+0,
-     -4.5353562730201404017e+0,
-      7.4819657596286408905e+0,
-     -6.0516124315132409155e+0,
-      2.4733515209909815443e+0,
-     -4.6937565143754629578e-1,
-      3.1608910440687221695e-2,
-     -2.4630612614645039828e-4
+      0.9999999999999999502e+0,
+     -2.6883926818565423430e+0,
+      2.6477222699473109692e+0,
+     -1.1538559607887416355e+0,
+      2.0886077795020607837e-1,
+     -1.0859777134152463084e-2
    };
    const double Q[] = {
       1.0000000000000000000e+0,
-     -4.5355682121856044935e+0,
-      8.1790029773247428573e+0,
-     -7.4634190853767468810e+0,
-      3.6245392503925290187e+0,
-     -8.9936784740041174897e-1,
-      9.8554565816757007266e-2,
-     -3.2116618742475189569e-3
+     -2.9383926818565635485e+0,
+      3.2712093293018635389e+0,
+     -1.7076702173954289421e+0,
+      4.1596017228400603836e-1,
+     -3.9801343754084482956e-2,
+      8.2743668974466659035e-4
    };
 
    double y = 0, r = 0, s = 1;
@@ -63,17 +60,14 @@ double li2(double x)
       s = -1;
    }
 
-   const double z = y - 0.25;
-   const double z2 = z*z;
-   const double z4 = z2*z2;
-   const double p = P[0] + z * P[1]
-      + z2 * (P[2] + z * P[3])
-      + z4 * (P[4] + z * P[5]
-              + z2 * (P[6] + z * P[7]));
-   const double q = Q[0] + z * Q[1]
-      + z2 * (Q[2] + z * Q[3])
-      + z4 * (Q[4] + z * Q[5]
-              + z2 * (Q[6] + z * Q[7]));
+   const double y2 = y*y;
+   const double y4 = y2*y2;
+   const double p = P[0] + y * P[1]
+      + y2 * (P[2] + y * P[3])
+      + y4 * (P[4] + y * P[5]);
+   const double q = Q[0] + y * Q[1]
+      + y2 * (Q[2] + y * Q[3])
+      + y4 * (Q[4] + y * Q[5] + y2 * Q[6]);
 
    return r + s*y*p/q;
 }

### 2021-12-12 anc/Li2.c
remove comment line

@@ -23,7 +23,6 @@ double li2(double x)
 
    double y = 0, r = 0, s = 1;
 
-   /* transform to [0, 1/2] */
    if (x < -1) {
       const double l = log(1 - x);
       y = 1/(1 - x);

### 2021-12-13 paper.tex
adding infty symbols
don't use widetext
which cuts of text
shrink equation

@@
 \usepackage{booktabs}
-\usepackage{widetext}
 \usepackage{hyperref}
@@
 be mapped onto the interval $[0,1/2]$ by using the following
-identities:
+identities \cite{lewin}:
 %
-\begin{widetext}%
 \begin{align}%
-  \Li(x) &= \Li\left(\frac{1}{1-x}\right) - \frac{\pi^2}{6} + \ln(1-x)\left[\frac{1}{2}\ln(1-x) - \ln(-x)\right], && (x<-1), \\
-  \Li(x) &= -\Li\left(\frac{x}{x-1}\right) - \frac{1}{2}\ln^2(1-x), && (-1<x<0), \\
-  \Li(x) &= -\Li(1-x) + \frac{\pi^2}{6} - \ln(x) \ln(1-x), && (\frac{1}{2}\leq x<1), \\
-  \Li(x) &= \Li\left(1-\frac{1}{x}\right) + \frac{\pi^2}{6} - \ln(x)\left[\ln\left(1-\frac{1}{x}\right) + \frac{1}{2}\ln(x)\right], && (1<x<2), \\
-  \Li(x) &= -\Li\left(\frac{1}{x}\right) + \frac{\pi^2}{3} - \frac{1}{2} \ln^2(x), && (x\geq 2),%
+  \Li(x) ={}& \Li\left(\frac{1}{1-x}\right) - \frac{\pi^2}{6} \notag \\ &+ \ln(1-x)\left[\frac{1}{2}\ln(1-x) - \ln(-x)\right], \\ % (x<-1), \\
+  \Li(x) ={}& -\Li\left(\frac{x}{x-1}\right) - \frac{1}{2}\ln^2(1-x), \\ % (-1<x<0), \\
+  \Li(x) ={}& -\Li(1-x) + \frac{\pi^2}{6} - \ln(x) \ln(1-x), \\ % (\frac{1}{2}\leq x<1), \\
+  \Li(x) ={}& \Li\left(1-\frac{1}{x}\right) + \frac{\pi^2}{6} \notag \\ &- \ln(x)\left[\ln\left(1-\frac{1}{x}\right) + \frac{1}{2}\ln(x)\right], \\ % (1<x<2), \\
+  \Li(x) ={}& -\Li\left(\frac{1}{x}\right) + \frac{\pi^2}{3} - \frac{1}{2} \ln^2(x), % (x\geq 2),%
 \end{align}%
-\end{widetext}%
 %
-as well as the special values $\Li(0)=0$, $\Li(1)=\pi^2/6$ and
-$\Li(-1)=-\pi^2/12$.  Within the interval $[0,1/2]$ the real
-dilogarithm has the series expansion \eqref{eq:series}, which can be
-written as
+and by using the special values $\Li(-1)=-\pi^2/12$, $\Li(0)=0$,
+$\Li(1/2)=\pi^2/12-\ln^2(2)/2$, $\Li(1)=\pi^2/6$ and $\Li(2)=\pi^2/4$.
+Within the interval $[0,1/2]$ the real dilogarithm has the series
+expansion \eqref{eq:series}, which can be written as
 %
 \begin{align}
-  \Li(x) = x \sum_{k=0} \frac{x^{k}}{(k+1)^2}.
+  \Li(x) = x \sum_{k=0}^\infty \frac{x^{k}}{(k+1)^2}.
   \label{eq:series2}
@@
 \begin{align}
-  \sum_{k=0} \frac{x^{k}}{(k+1)^2} =
+  \sum_{k=0}^\infty \frac{x^{k}}{(k+1)^2} \approx
   \frac{\sum_{k=0}^5 p_kx^k}{\sum_{k=0}^6 q_kx^k}.

### 2021-12-14 paper.tex
adding ranges
refine ranges
refine range
remove parenthesis
shuffle terms
improve formulation

@@
   \Li(x) &= -\int_0^x \frac{\ln(1-t)}{t}dt, &
-  &(x<1).
+  &x<1.
 \end{align}
@@
 %
-\begin{align}%
-  \Li(x) ={}& \Li\left(\frac{1}{1-x}\right) - \frac{\pi^2}{6} \notag \\ &+ \ln(1-x)\left[\frac{1}{2}\ln(1-x) - \ln(-x)\right], \\ % (x<-1), \\
-  \Li(x) ={}& -\Li\left(\frac{x}{x-1}\right) - \frac{1}{2}\ln^2(1-x), \\ % (-1<x<0), \\
-  \Li(x) ={}& -\Li(1-x) + \frac{\pi^2}{6} - \ln(x) \ln(1-x), \\ % (\frac{1}{2}\leq x<1), \\
-  \Li(x) ={}& \Li\left(1-\frac{1}{x}\right) + \frac{\pi^2}{6} \notag \\ &- \ln(x)\left[\ln\left(1-\frac{1}{x}\right) + \frac{1}{2}\ln(x)\right], \\ % (1<x<2), \\
-  \Li(x) ={}& -\Li\left(\frac{1}{x}\right) + \frac{\pi^2}{3} - \frac{1}{2} \ln^2(x), % (x\geq 2),%
-\end{align}%
+\begin{align}
+  \Li(x) ={}& \ln(1-x)\left[\frac{1}{2}\ln(1-x) - \ln(-x)\right] - \frac{\pi^2}{6}
+              \notag \\ &+\Li\left(\frac{1}{1-x}\right), \quad x\leq -1, \\
+  \Li(x) ={}& -\Li\left(\frac{x}{x-1}\right) - \frac{1}{2}\ln^2(1-x),
+              \notag \\ & -1<x<0, \\
+  \Li(x) ={}& -\Li(1-x) + \frac{\pi^2}{6} - \ln(x) \ln(1-x),
+              \notag \\ &\frac{1}{2}<x<1, \\
+  \Li(x) ={}& \frac{\pi^2}{6} - \ln(x)\left[\ln\left(1-\frac{1}{x}\right) + \frac{1}{2}\ln(x)\right]
+              \notag \\ &+ \Li\left(1-\frac{1}{x}\right), \quad 1<x\leq 2, \\
+  \Li(x) ={}& -\Li\left(\frac{1}{x}\right) + \frac{\pi^2}{3} - \frac{1}{2} \ln^2(x), \quad x>2,
+\end{align}
 %
@@
 %
-We approximate the sum on the r.h.s.\ of Eq.~\eqref{eq:series2} using
-a rational minimax approximant on the interval $[0,1/2]$ using the
-\texttt{MiniMaxApproximation} from Wolfram/Mathematica
+We approximate the sum on the r.h.s.\ of Eq.~\eqref{eq:series2} in the
+form of a rational minimax approximant on the interval $[0,1/2]$ using
+the \texttt{MiniMaxApproximation} from Wolfram/Mathematica
 \cite{mathematica},

### 2021-12-14 paper.tex
use other i7 machine

@@
 compared on two different CPU architectures.  On each architecture the
-implentations are compiled with \texttt{g++} 10.2.1 with optimization
+implentations are compiled with \texttt{g++} with optimization
 \texttt{-O2}.  The shown runtime denotes the total wall clock time for
@@
   \caption{Runtime in seconds for the calculation of $\Li(x)$ for
-    different implementations on different CPU architectures.}
+    different implementations on different CPU architectures with
+    different compilers.}
   \begin{tabular}{lcc}
     \toprule
-    Implementation & i5-8265U & i7-5600U \\
+    & i7-4700MQ & i7-5600U \\
+    Implementation & \texttt{g++} 10.2.1 & \texttt{g++} 9.3.0 \\
     \midrule
-    this paper & $0.022$ & $0.026$ \\
-    Morris (DILOG 0011) \cite{morris} & $0.022$ & $0.026$ \\
-    Cephes \cite{cephes} & $0.028$ & $0.034$ \\
-    Algorithm 327 \cite{koelbigDilog} & $0.031$ & $0.035$ \\
-    Chebyshev \cite{luke} & $0.058$ & $0.060$ \\
-    Algorithm 490 \cite{ginsberg} & $0.059$ & $0.071$ \\
-    GSL \cite{gsl} & & $0.155$ \\
+    this paper & $0.031$ & $0.026$ \\
+    Morris (DILOG 0011) \cite{morris} & $0.035$ & $0.026$ \\
+    Cephes \cite{cephes} & $0.044$ & $0.034$ \\
+    Algorithm 327 \cite{koelbigDilog} & $0.054$ & $0.035$ \\
+    Chebyshev \cite{luke} & $0.079$ & $0.060$ \\
+    Algorithm 490 \cite{ginsberg} & $0.125$ & $0.071$ \\
+    GSL \cite{gsl} & $0.330$ & $0.155$ \\
     \bottomrule

### 2021-12-14 paper.tex
improve table head
extend abstract
reduce strictness

@@
 \usepackage{hyperref}
+\usepackage{multirow}
 \setkomafont{disposition}{\bfseries}
@@
 An algorithm for the computation of the real dilogarithm is presented,
-which allows for the use of instruction\-/level parallelism.
+which allows for the use of instruction\-/level parallelism when
+implemented on appropriate CPUs.
 
@@
 calculation of the real dilogarithm $\Li$ is presented, which makes
-use of ILP, resulting in a reduced runtime compared to the algorithms
-presented so far.
+use of ILP, potentially resulting in a reduced runtime compared to
+most algorithms presented so far.
 
@@
     \toprule
-    & i7-4700MQ & i7-5600U \\
-    Implementation & \texttt{g++} 10.2.1 & \texttt{g++} 9.3.0 \\
+    \multirow{2}{*}{Implementation} & i7-4700MQ & i7-5600U \\
+    & \texttt{g++} 10.2.1 & \texttt{g++} 9.3.0 \\
     \midrule

### 2021-12-14 paper.tex
improve formulation

@@
 %
-around $x=0$ for $x\in(-1,1]$, where terms of $O(x^8)$ and beyond have
-been omitted.  In the following we will briefly show various possible
-implementations of this function and discuss the performance of the
-different implementations.  For brevity we will restrict ourselves to
-$x\in(-1,1]$.  In C the function in Eq.~\eqref{eq:taylor} could be
-implemented very naively as follows:
+around $x=0$ for $x\in(-1,1]$.  In Eq.~\eqref{eq:taylor} terms of
+$O(x^8)$ and beyond have been omitted for brevity.  In the following
+we will briefly show various possible implementations of this function
+and discuss the performance of the different implementations.  For
+brevity we will restrict ourselves to $x\in(-1,1]$.  In C the function
+in Eq.~\eqref{eq:taylor} could be implemented very naively as follows:
 %
@@
   rewrite expensive floating\-/point divisions by a constant, such as
-  $x/2$, as less expensive floating\-/point multiplication by a
+  $x/2$, as less expensive floating\-/point multiplications by a
   constant, for example as $0.5x$.  In the following we will therefore
-  count a division of a variable by a constant as a floating\-/point
-  multiplication.}
+  count a division of a floating\-/point variable by a constant as a
+  floating\-/point multiplication.}
 
@@
 \lstinputlisting[language=C]{src/log1p_horner.c}
+%
+Note, that the formulation using Horner's scheme requires the
+evaluation of the floating\-/point operations in strictly sequential
+order.
 
@@
 polynomials, where the cost of the floating\-/point division is
-smaller than the gain by the parallel execution.
+smaller than the gain by the parallel execution of the terms in the
+numerator and denominator.
 

### 2021-12-15 paper.tex
improve formulation
refine abstact and title

@@
 
-\newcommand{\mytitle}{An algorithm for the calculation of the real dilogarithm with instruction-level parallelism}
+\newcommand{\mytitle}{Comparisons of methods for the calculation of the real dilogarithm regarding instruction-level parallelism}
 \newcommand{\myauthor}{Alexander Voigt}
@@
 
-An algorithm for the computation of the real dilogarithm is presented,
-which allows for the use of instruction\-/level parallelism when
-implemented on appropriate CPUs.
+We compare different methods for the computation of the real
+dilogarithm regarding their ability for using instruction\-/level
+parallelism when executed on appropriate CPUs.  As a result we present
+an instruction\-/level\-/aware method and compare it to existing
+implementations.
 
@@
 %
-We find that the implementation presented in this paper is fastest
-among the ones shown in the table.  The next-to fastest implementation
-is provided by Cephes, which is only $\sim 20\%$ slower.  The short
-runtime of Cephes stems from the fact that Cephes also uses a rational
-function approximation, where the numerator and the denominator can be
+We find that the implementation presented in this paper and the
+approach presented in \cite{morris} are fastest among the ones shown
+in the table.  The next-to fastest implementation is provided by
+Cephes, which is only $\sim 20\%$ slower.  The short runtime of Cephes
+stems from the fact that Cephes also uses a rational function
+approximation, where the numerator and the denominator can be
 evaluated in parallel.  However, Cephes evaluates the polynomials in

### 2021-12-15 paper.tex
avoid fp division

@@
 This naive implementation requires 26 floating\-/point multiplications
-and 6 floating\-/point additions.\footnote{Modern compilers usually
-  rewrite expensive floating\-/point divisions by a constant, such as
-  $x/2$, as less expensive floating\-/point multiplications by a
-  constant, for example as $0.5x$.  In the following we will therefore
-  count a division of a floating\-/point variable by a constant as a
-  floating\-/point multiplication.}
+and 6 floating\-/point additions.
 

### 2021-12-15 src/log1p_estrin.c

@@ -1,6 +1,6 @@
 double log1p_estrin(double x) {
    const double y = x*x;
    const double z = y*y;
-   return x + (x/3 - 0.5)*y
-      + (y*(-1./6 + x/7) + x/5 - 0.25)*z;
+   return x + (1./3*x - 1./2)*y
+      + (y*(-1./6 + 1./7*x) + x/5 - 1./4)*z;
 }

### 2021-12-15 src/log1p_horner.c

@@ -1,5 +1,5 @@
 double log1p_horner(double x) {
-   return x*(1 + x*(-0.5 + x*(1./3
-     + x*(-0.25 + x*(0.2
-     + x*(-1./6 + x/7))))));
+   return x*(1 + x*(-1./2 + x*(1./3
+     + x*(-1./4 + x*(1./5
+     + x*(-1./6 + 1./7*x))))));
 }

### 2021-12-15 src/log1p_naive.c

@@ -1,5 +1,5 @@
 double log1p_naive(double x) {
-   return x - x*x/2 + x*x*x/3 - x*x*x*x/4
-      + x*x*x*x*x/5 - x*x*x*x*x*x/6
-      + x*x*x*x*x*x*x/7;
+   return x - 1./2*x*x + 1./3*x*x*x
+      - 1./4*x*x*x*x + 1./5*x*x*x*x*x
+      - 1./6*x*x*x*x*x*x + 1./7*x*x*x*x*x*x*x;
 }

### 2021-12-15 src/log1p_estrin.c
avoid another fp division

@@ -1,6 +1,6 @@
 double log1p_estrin(double x) {
    const double y = x*x;
    const double z = y*y;
-   return x + (1./3*x - 1./2)*y
-      + (y*(-1./6 + 1./7*x) + x/5 - 1./4)*z;
+   return x + (1./3*x - 1./2)*y +
+      (y*(-1./6 + 1./7*x) + 1./5*x - 1./4)*z;
 }

### 2021-12-15 paper.tex
slight reformulation of the intro

@@
 
-The development of modern CPUs with support for instruction\-/level
+Many of the publicly available implementations, e.g.\ \cite{root,gsl},
+use formulations for the approximation of the dilogarithm with a
+strictly sequential execution of floating\-/point operations.  The
+development of modern CPUs with support for instruction\-/level
 parallelism (ILP) opens the opportunity for a further reduction of the
@@
 appropriately in order to allow for the parallel execution of multiple
-floating\-/point operations.  In this paper an algorithm for the
+floating\-/point operations.  In this paper a method for the
 calculation of the real dilogarithm $\Li$ is presented, which makes
 use of ILP, potentially resulting in a reduced runtime compared to
-most algorithms presented so far.
+many algorithms presented so far.
 
@@
 implementation strategies are discussed using a toy example function.
-In \secref{sec:algorithm} an ILP-aware algorithm for the numerical
+In \secref{sec:algorithm} an ILP-aware method for the numerical
 calculation of the dilogarithm is presented.  A C implementation of
-this algorithm can be found in the appendix and in the arXiv
-submission of this paper.
+this method can be found in the appendix and in the arXiv submission
+of this paper.
 
@@
 
-\section{An ILP-aware algorithm for the real dilogarithm}
+\section{An ILP-aware method for the real dilogarithm}
 \label{sec:algorithm}

### 2021-12-16 paper.tex
reformulation

@@
 %
-This naive implementation requires 26 floating\-/point multiplications
-and 6 floating\-/point additions.
+This naive implementation requires 27 floating\-/point multiplications
+and 6 floating\-/point additions.  Although all summands in this
+implementation can in principle be calculated in parallel, the cost of
+the many (redundant) floating\-/point multiplications is usually so
+high that this formulated is in total very time\-/inefficient.
 
@@
 evaluation of the floating\-/point operations in strictly sequential
-order.
+order.  However, the significant reduction of the number of
+floating\-/point multiplications, compared to the naive
+implementation, usually results in a shorter runtime.
 

### 2021-12-16 paper.tex
reformulation

@@
 
-Several modern CPU architectures support instruction\-/level
-parallelism (ILP), which allows for the execution of multiple
-instructions at the same time.  To make use of ILP, one could use
-Estrin's scheme \cite{estrin} to rewrite the function as
+There are several different possibilities to reduce the runtime of
+Horner's method by rewriting the series expansion of $\ln(1+x)$ such
+that the more terms can be calculated in parallel, while the number
+floating\-/point operations is still kept small.  One possibility is
+to split the r.h.s.\ of Eq.~\eqref{eq:horner} into two distinct sums,
+one containing the even powers and the other containing the odd powers
+of $x$.  Both sums can then be calculated in parallel.  Another
+possibility is to rewrite r.h.s.\ of Eq.~\eqref{eq:taylor} using
+Estrin's scheme \cite{estrin},
 %
@@
 where $y=x^2$ and $z=y^2$.  This form requires 8 floating\-/point
-multiplications and 6 additions, i.e.\ more multiplications than
+multiplications and 6 additions, i.e.\ one more multiplication than
 Horner's scheme.  However, in Estrin's form several terms can be

### 2021-12-16 paper.tex
extend text

@@
 one containing the even powers and the other containing the odd powers
-of $x$.  Both sums can then be calculated in parallel.  Another
-possibility is to rewrite r.h.s.\ of Eq.~\eqref{eq:taylor} using
-Estrin's scheme \cite{estrin},
+of $x$,
+%
+\begin{align}
+\begin{split}
+  \ln(1+x) \approx{}& x \Bigg\{1 + y \Bigg[\frac{1}{3} + y
+  \Bigg(\frac{1}{5} + \frac{y}{7}\Bigg)\Bigg]\Bigg\} \\
+  &+ y\Bigg[-\frac{1}{2} + y\Bigg(-\frac{1}{4} - \frac{y}{6}\Bigg)\Bigg],
+\end{split}\label{eq:split}%
+\end{align}
+%
+where $y=x^2$.  This formulation requires 8 floating\-/point
+multiplications and 6 additions, i.e.\ one more multiplication than
+Horner's scheme.  However, in Eq.~\eqref{eq:split} both sums can then
+be calculated in parallel, leading to a potential speed-up if the cost
+of the additional multiplication is smaller than the gain by the
+parallel execution.  A C implementation could read:
+%
+\lstinputlisting[language=C]{src/log1p_split.c}
+%
+Another possibility is to rewrite the r.h.s.\ of Eq.~\eqref{eq:taylor}
+using Estrin's scheme \cite{estrin},
 %
@@
 %
-where $y=x^2$ and $z=y^2$.  This form requires 8 floating\-/point
-multiplications and 6 additions, i.e.\ one more multiplication than
-Horner's scheme.  However, in Estrin's form several terms can be
-executed in parallel, leading to a potential speed-up if the cost of
-the additional multiplications is smaller than the gain by the
-parallel execution.  A C implementation using Estrin's scheme could
-read:
+where $y=x^2$ and $z=y^2$.  This form also requires 8 floating\-/point
+multiplications and 6 additions, as in Eq.~\eqref{eq:split}.  However,
+in Estrin's form more terms can be executed in parallel, leading to a
+potential further speed-up.  A C implementation using Estrin's scheme
+could read:
 %

### 2021-12-16 src/log1p.h

@@ -1,5 +1,6 @@
 double log1p_naive(double);
 double log1p_horner(double);
+double log1p_split(double);
 double log1p_estrin(double);
 double log1p_pade(double);
 double log1p_mixed(double);
@@ -12,6 +13,7 @@ struct Fn {
 const struct Fn functions[] = {
    { log1p_naive , "log1p_naive"  },
    { log1p_horner, "log1p_horner" },
+   { log1p_split , "log1p_split"  },
    { log1p_estrin, "log1p_estrin" },
    { log1p_pade  , "log1p_pade"   },
    { log1p_mixed , "log1p_mixed"  },

### 2021-12-16 src/log1p_split.c

@@ -1,0 +1,5 @@
+double log1p_split(double x) {
+   const double y = x*x;
+   return x*(1 + y*(1./3 + y*(1./5 + 1./7*y)))
+      + y*(-1./2 + y*(-1./4 - 1./6*y));
+}

### 2021-12-16 src/log1p_split.c
simplify equations

@@ -1,5 +1,5 @@
 double log1p_split(double x) {
    const double y = x*x;
    return x*(1 + y*(1./3 + y*(1./5 + 1./7*y)))
-      + y*(-1./2 + y*(-1./4 - 1./6*y));
+      - y*(1./2 + y*(1./4 + 1./6*y));
 }

### 2021-12-16 paper.tex
reduce bracket sizes
simplify equations
use interval [0,1/2]

@@
 \begin{split}
-  \ln(1+x) \approx{}& x \Bigg\{1 + x \Bigg[-\frac{1}{2} + x \Bigg(\frac{1}{3} + x \Bigg\{-\frac{1}{4} \\
-  &+ x \Bigg[\frac{1}{5} + x\Bigg(-\frac{1}{6} + \frac{x}{7}\Bigg)\Bigg]\Bigg\}\Bigg)\Bigg]\Bigg\}.
+  \ln(1+x) \approx{}& x \bigg\{1 + x \bigg[-\frac{1}{2} + x \bigg(\frac{1}{3} + x \bigg\{-\frac{1}{4} \\
+  &+ x \bigg[\frac{1}{5} + x\bigg(-\frac{1}{6} + \frac{x}{7}\bigg)\bigg]\bigg\}\bigg)\bigg]\bigg\}.
 \end{split}\label{eq:horner}%
@@
 \begin{split}
-  \ln(1+x) \approx{}& x \Bigg\{1 + y \Bigg[\frac{1}{3} + y
-  \Bigg(\frac{1}{5} + \frac{y}{7}\Bigg)\Bigg]\Bigg\} \\
-  &+ y\Bigg[-\frac{1}{2} + y\Bigg(-\frac{1}{4} - \frac{y}{6}\Bigg)\Bigg],
+  \ln(1+x) \approx{}& x \left\{1 + y \left[\frac{1}{3} + y
+  \left(\frac{1}{5} + \frac{y}{7}\right)\right]\right\} \\
+  &- y\left[\frac{1}{2} + y\left(\frac{1}{4} + \frac{y}{6}\right)\right],
 \end{split}\label{eq:split}%
@@
 the calculation of $\Li(x)$ for $10^6$ uniformly distributed random
-values $x\in[-5,5]$.
+values $x\in[0,1/2]$.
 %
@@
     \toprule
-    \multirow{2}{*}{Implementation} & i7-4700MQ & i7-5600U \\
+    \multirow{2}{*}{Implementation} & i7-5600U & ?? \\
     & \texttt{g++} 10.2.1 & \texttt{g++} 9.3.0 \\
     \midrule
-    this paper & $0.031$ & $0.026$ \\
-    Morris (DILOG 0011) \cite{morris} & $0.035$ & $0.026$ \\
-    Cephes \cite{cephes} & $0.044$ & $0.034$ \\
-    Algorithm 327 \cite{koelbigDilog} & $0.054$ & $0.035$ \\
-    Chebyshev \cite{luke} & $0.079$ & $0.060$ \\
-    Algorithm 490 \cite{ginsberg} & $0.125$ & $0.071$ \\
-    GSL \cite{gsl} & $0.330$ & $0.155$ \\
+    this paper                        & $0.0058$ & $??$ \\
+    Morris (DILOG 0011) \cite{morris} & $0.0085$ & $??$ \\
+    Cephes \cite{cephes}              & $0.0119$ & $??$ \\
+    Algorithm 327 \cite{koelbigDilog} & $0.0172$ & $??$ \\
+    Chebyshev \cite{luke}             & $0.0515$ & $??$ \\
+    Algorithm 490 \cite{ginsberg}     & $0.0576$ & $??$ \\
+    GSL \cite{gsl}                    & $0.0898$ & $??$ \\
     \bottomrule

### 2021-12-16 paper.tex
update numbers

@@
 the calculation of $\Li(x)$ for $10^6$ uniformly distributed random
-values $x\in[0,1/2]$.
+values $x\in[0,1/2]$.  The choice of this interval ensures that the
+measurement is only sensitive to the formulation of the approximant,
+but not to the transformation onto the interval $[0,1/2]$.
 %
@@
     \toprule
-    \multirow{2}{*}{Implementation} & i7-5600U & ?? \\
-    & \texttt{g++} 10.2.1 & \texttt{g++} 9.3.0 \\
+    \multirow{2}{*}{Implementation} & i7-4700MQ & i7-5600U \\
+    & \texttt{g++} 9.3.0 & \texttt{g++} 10.2.1 \\
     \midrule
-    this paper                        & $0.0058$ & $??$ \\
-    Morris (DILOG 0011) \cite{morris} & $0.0085$ & $??$ \\
-    Cephes \cite{cephes}              & $0.0119$ & $??$ \\
-    Algorithm 327 \cite{koelbigDilog} & $0.0172$ & $??$ \\
-    Chebyshev \cite{luke}             & $0.0515$ & $??$ \\
-    Algorithm 490 \cite{ginsberg}     & $0.0576$ & $??$ \\
-    GSL \cite{gsl}                    & $0.0898$ & $??$ \\
+    this paper                        & $0.0076$ & $0.0058$ \\
+    Morris (DILOG 0011) \cite{morris} & $0.0115$ & $0.0085$ \\
+    Cephes \cite{cephes}              & $0.0136$ & $0.0119$ \\
+    Algorithm 327 \cite{koelbigDilog} & $0.0229$ & $0.0172$ \\
+    Chebyshev \cite{luke}             & $0.0718$ & $0.0515$ \\
+    Algorithm 490 \cite{ginsberg}     & $0.1057$ & $0.0576$ \\
+    GSL \cite{gsl}                    & $0.2091$ & $0.0898$ \\
     \bottomrule

### 2021-12-16 paper.tex
remove intermediate implementation

@@
 \begin{align}
-  \ln(1+x) \approx x
+  \ln(1+x) &\approx x
   \frac{\frac{320 x^3}{40143}+\frac{44320 x^2}{120429}+\frac{51064
     x}{40143}+1}{\frac{5120 x^3}{40143}+\frac{12320 x^2}{13381}+\frac{23712
-    x}{13381}+1}.
+    x}{13381}+1}
   \label{eq:pade}
-\end{align}
-%
-A naive C implementation of Eq.~\eqref{eq:pade} could read:
-%
-\lstinputlisting[language=C]{src/log1p_pade.c}
-%
-The numerator and the denominator of Eq.~\eqref{eq:pade} can be
-re-written using Estrin's scheme, allowing for further ILP:
-%
-\begin{align}
-  \ln(1+x) \approx x
+  \\
+  &= x
   \frac{\left(\frac{320 x}{40143}+\frac{44320}{120429}\right)y+\frac{51064
@@
 %
-where $y=x^2$.  A C implementation could read:
+where $y=x^2$.  In Eq.~\eqref{eq:mixed} the numerator and the
+denominator have been re-written using Estrin's scheme, allowing for
+further ILP.  A C implementation of Eq.~\eqref{eq:mixed} could read:
 %

### 2021-12-18 paper.tex
typo

@@
 Horner's method by rewriting the series expansion of $\ln(1+x)$ such
-that the more terms can be calculated in parallel, while the number
+that more terms can be calculated in parallel, while the number
 floating\-/point operations is still kept small.  One possibility is

### 2021-12-18 paper.tex
extend text

@@
 The coefficients $p_k$ and $q_k$ are listed in \tabref{tab:coeffs}.
+The approximant on the r.h.s.\ of Eq.~\eqref{eq:minimax} has a maximum
+relative error of around $5\cdot 10^{-17}$, suited for an evaluation
+of $\Li(x)$ with double precision.
 %
@@
 \end{table}
+%
+Note that the approximant on the r.h.s.\ of Eq.~\eqref{eq:minimax} is
+very similar to the results presented in Ref.~\cite{morris} (DILOG
+0011), although with slightly different coefficients.
 

### 2021-12-19 paper.tex
extend text

@@
 very similar to the results presented in Ref.~\cite{morris} (DILOG
-0011), although with slightly different coefficients.
+0011), although with slightly different coefficients.  In oder to
+allow for the use of ILP for the evaluation of the numerator and the
+denominator we use Estrin's scheme,
+%
+\begin{align}
+  \Li(x) &\approx x\frac{P(x)}{Q(x)}, \qquad 0\leq x \leq \frac{1}{2}, \label{eq:final}\\
+  P(x) &= p_0 + x p_1 + y (p_2 + x p_3) + z (p_4 + x p_5), \\
+  Q(x) &= q_0 + x q_1 + y (q_2 + x q_3) + z (q_4 + x q_5 + y q_6),
+\end{align}
+%
+with $y=x^2$ and $z=y^2$.  A C implementation of Eq.~\eqref{eq:final}
+can be found in the appendix.
 
@@
     \midrule
-    this paper                        & $0.0076$ & $0.0058$ \\
+    Eq.~\eqref{eq:final} (this paper) & $0.0076$ & $0.0058$ \\
     Morris (DILOG 0011) \cite{morris} & $0.0115$ & $0.0085$ \\
@@
 %
-We find that the implementation presented in this paper and the
-approach presented in \cite{morris} are fastest among the ones shown
-in the table.  The next-to fastest implementation is provided by
-Cephes, which is only $\sim 20\%$ slower.  The short runtime of Cephes
-stems from the fact that Cephes also uses a rational function
-approximation, where the numerator and the denominator can be
-evaluated in parallel.  However, Cephes evaluates the polynomials in
-the numerator and denominator using Horner's scheme, which is purely
-sequential.  Furthermore, for $x>1$ Cephes calculates $\ln(x)$ twice,
-which leads to a further slowdown.
+We find that the implementation presented in Eq.~\eqref{eq:final} in
+this paper and the approach presented in Ref.~\cite{morris} (DILOG
+0011) are fastest among the ones shown in the table.  Both approaches
+use a rational function approximant with the same number of terms.
+However, the approximant from Ref.~\cite{morris} has been implemented
+(for comparison) to use Horner's scheme to evaluate the numerator and
+the denominator.  Thus, the runtime difference of $\sim 30\%$ between
+these two stems only from the fact that the implementation presented
+in this paper uses Estrin's scheme to evaluate the numerator and the
+denominator and as such can make use of ILP, while in the
+implementation of Ref.~\cite{morris} the numerator and the denominator
+must be evaluated sequentially.
+
+The third fastest implementation is provided by Cephes, which is by a
+factor two slower than the implementation in the appendix.  This
+longer runtime has several reasons: For one, Cephes uses a rational
+function approximation with more terms in the numerator and
+denominator.  Furthermore, Cephes evaluates the numerator and
+denominator polynomials using Horner's scheme, which is purely
+sequential.\footnote{For $x>1$ Cephes calculates $\ln(x)$ twice, which
+  leads to a further increased runtime.}
 

### 2021-12-19 paper.tex
extend footnode

@@
 sequential.\footnote{For $x>1$ Cephes calculates $\ln(x)$ twice, which
-  leads to a further increased runtime.}
+  leads to a further increased runtime when a different interval for
+  $x$ is chosen.}
 

### 2021-12-19 paper.tex
split equation

@@
 \begin{align}
-  \Li(x) &\approx x\frac{P(x)}{Q(x)}, \qquad 0\leq x \leq \frac{1}{2}, \label{eq:final}\\
+  \Li(x) &\approx x\frac{P(x)}{Q(x)}, \qquad 0\leq x \leq \frac{1}{2},
+  \label{eq:final}
+\end{align}
+%
+with
+%
+\begin{align}
   P(x) &= p_0 + x p_1 + y (p_2 + x p_3) + z (p_4 + x p_5), \\
@@
 %
-with $y=x^2$ and $z=y^2$.  A C implementation of Eq.~\eqref{eq:final}
+where $y=x^2$ and $z=y^2$.  A C implementation of Eq.~\eqref{eq:final}
 can be found in the appendix.
 
+\section{Runtime comparison}
+
 In \tabref{tab:runtime} the runtime of different C implementations is
-compared on two different CPU architectures.  On each architecture the
-implentations are compiled with \texttt{g++} with optimization
-\texttt{-O2}.  The shown runtime denotes the total wall clock time for
-the calculation of $\Li(x)$ for $10^6$ uniformly distributed random
-values $x\in[0,1/2]$.  The choice of this interval ensures that the
-measurement is only sensitive to the formulation of the approximant,
-but not to the transformation onto the interval $[0,1/2]$.
+compared to the implementation given in the appednix on two different
+CPU architectures.  On each architecture the implentations are
+compiled with \texttt{g++} with optimization \texttt{-O2}.  The shown
+runtime denotes the total wall clock time for the calculation of
+$\Li(x)$ for $10^6$ uniformly distributed random values $x\in[0,1/2]$.
+The choice of this interval ensures that the measurement is only
+sensitive to the formulation of the approximant, but not to the
+transformation onto the interval $[0,1/2]$.
 %

### 2021-12-19 paper.tex
improve formulation

@@
 sensitive to the formulation of the approximant, but not to the
-transformation onto the interval $[0,1/2]$.
+transformation onto the interval $[0,1/2]$.%
 %
@@
   \label{tab:runtime}
-\end{table}
-%
-We find that the implementation presented in Eq.~\eqref{eq:final} in
-this paper and the approach presented in Ref.~\cite{morris} (DILOG
-0011) are fastest among the ones shown in the table.  Both approaches
-use a rational function approximant with the same number of terms.
-However, the approximant from Ref.~\cite{morris} has been implemented
-(for comparison) to use Horner's scheme to evaluate the numerator and
-the denominator.  Thus, the runtime difference of $\sim 30\%$ between
+\end{table}%
+
+We find that the implementation presented in this paper and the
+approach presented in Ref.~\cite{morris} (DILOG 0011) are fastest
+among the ones shown in the table.  Both approaches use a rational
+function approximant with the same number of terms.  However, the
+approximant from Ref.~\cite{morris} has been implemented (for
+comparison) to use Horner's scheme to evaluate the numerator and the
+denominator.  Thus, the runtime difference of $\sim 30\%$ between
 these two stems only from the fact that the implementation presented

### 2021-12-19 paper.tex
formulation
improve formulation

@@
 The third fastest implementation is provided by Cephes, which is by a
-factor two slower than the implementation in the appendix.  This
-longer runtime has several reasons: For one, Cephes uses a rational
-function approximation with more terms in the numerator and
+factor two slower than the implementation prestend in this paper.
+This longer runtime has several reasons: For one, Cephes uses a
+rational function approximation with more terms in the numerator and
 denominator.  Furthermore, Cephes evaluates the numerator and
@@
 sequential.\footnote{For $x>1$ Cephes calculates $\ln(x)$ twice, which
-  leads to a further increased runtime when a different interval for
-  $x$ is chosen.}
+  leads to a further increased runtime when a benchmark interval with
+  $x>1$ is chosen.}
 

### 2021-12-20 paper.tex
improve language

@@
 denominator.  Thus, the runtime difference of $\sim 30\%$ between
-these two stems only from the fact that the implementation presented
+the two stems only from the fact that the implementation presented
 in this paper uses Estrin's scheme to evaluate the numerator and the

### 2021-12-20 paper.tex
adding ref. to table

@@
     Algorithm 327 \cite{koelbigDilog} & $0.0229$ & $0.0172$ \\
-    Chebyshev \cite{luke}             & $0.0718$ & $0.0515$ \\
+    Chebyshev \cite{luke,root}        & $0.0718$ & $0.0515$ \\
     Algorithm 490 \cite{ginsberg}     & $0.1057$ & $0.0576$ \\

### 2021-12-20 paper.tex
extend text

@@
 
+Algorithm 327 \cite{koelbigDilog}, Algorithm 490 \cite{ginsberg} and
+the implementation found in the Gnu Scientific Library (GSL)
+\cite{gsl} use series expansions, which are summed using Horner's
+scheme.  These algorithms thus do not make use of ILP, which is one
+reason for their longer runtime.  Furthermore, Algorithm 490 and most
+series expansions in the GSL use floating\-/point divisions inside
+loops, which may be another reason for the increased runtime.
+
+% Chebyshev \cite{luke,root}
+
 \section{Conclusions}

### 2021-12-20 paper.tex
typo
extend text

@@
 Algorithm 327 \cite{koelbigDilog}, Algorithm 490 \cite{ginsberg} and
-the implementation found in the Gnu Scientific Library (GSL)
+the implementation found in the GNU Scientific Library (GSL)
 \cite{gsl} use series expansions, which are summed using Horner's
@@
 
-% Chebyshev \cite{luke,root}
+A representation in terms of Chebyshev polynomials \cite{luke} is
+implemented in the ROOT program library \cite{root}.  In this
+implementation the polynomials are summed using Clenshaw's algorithm,
+which is also purely sequencial and slightly more complex than
+Horner's scheme.  This is a reason why the implementation in
+\cite{root} has a larger runtime than the rational function
+approximants discussed above.
 

### 2021-12-20 paper.tex
extend conclusion

@@
 
-\newcommand{\mytitle}{Comparisons of methods for the calculation of the real dilogarithm regarding instruction-level parallelism}
+\newcommand{\mytitle}{Comparison of methods for the calculation of the real dilogarithm regarding instruction-level parallelism}
 \newcommand{\myauthor}{Alexander Voigt}
@@
 
-C, C++ and FORTRAN implementations of the presented algorithm can be
-found in the Polylogarithm library \cite{polylogarithm}.
-Implementations in Julia and Rust can be found in \cite{PolyLog.jl}
-and \cite{polylog}, respectively.
+In this paper different approaches for the numerical calculation of
+the real dilogarithm have been compared w.r.t.\ their potential for
+the use of instruction\-/level parallelism.  A particular
+implementation of a rational minimax approximant has been proposed and
+compared to several existing implementations w.r.t.\ runtime.  C, C++
+and FORTRAN implementations of the presented algorithm can be found in
+the Polylogarithm library \cite{polylogarithm}.  Implementations in
+Julia and Rust can be found in \cite{PolyLog.jl} and \cite{polylog},
+respectively.
 

### 2021-12-20 paper.tex
extend conclusions
move footnote down

@@
 Estrin's scheme, similar to the approach used in
-Eq.~\eqref{eq:mixed}.%
-\footnote{The Cephes library \cite{cephes} implements an equivalent
-  rational function approximation for Spence's function
-  $\operatorname{Sp}(x)=\Li(1-x)$.}
+Eq.~\eqref{eq:mixed}.
 
@@
 
-The third fastest implementation is provided by Cephes, which is by a
-factor two slower than the implementation prestend in this paper.
-This longer runtime has several reasons: For one, Cephes uses a
-rational function approximation with more terms in the numerator and
-denominator.  Furthermore, Cephes evaluates the numerator and
-denominator polynomials using Horner's scheme, which is purely
-sequential.\footnote{For $x>1$ Cephes calculates $\ln(x)$ twice, which
-  leads to a further increased runtime when a benchmark interval with
-  $x>1$ is chosen.}
+The third fastest implementation is provided by Cephes \cite{cephes},
+which implements a rational function approximation for Spence's
+function $\operatorname{Sp}(x)=\Li(1-x)$.  The implementation provided
+by Cephes is by a factor two slower than the implementation prestend
+in this paper.  This longer runtime has several reasons: For one,
+Cephes uses a rational function approximation with more terms in the
+numerator and denominator.  Furthermore, Cephes evaluates the
+numerator and denominator polynomials using Horner's scheme, which is
+purely sequential.\footnote{For $x>1$ Cephes calculates $\ln(x)$
+  twice, which leads to a further increased runtime when a benchmark
+  interval with $x>1$ is chosen.}
 
@@
 implementation of a rational minimax approximant has been proposed and
-compared to several existing implementations w.r.t.\ runtime.  C, C++
-and FORTRAN implementations of the presented algorithm can be found in
-the Polylogarithm library \cite{polylogarithm}.  Implementations in
-Julia and Rust can be found in \cite{PolyLog.jl} and \cite{polylog},
+compared to several existing implementations w.r.t.\ runtime.  A C
+implementations of the presented algorithm can be found in the
+appendix.  C++ and FORTRAN implementations can be found in the
+Polylogarithm library \cite{polylogarithm}.  Implementations in Julia
+and Rust can be found in \cite{PolyLog.jl} and \cite{polylog},
 respectively.

### 2021-12-20 paper.tex
rename sections
rename section
extend summary

@@
 
-\section{Comparison of implementation strategies}
+\section{Implementation strategies}
 \label{sec:toymodel}
@@
 
-\section{An ILP-aware method for the real dilogarithm}
+\section{ILP-aware approximation of the real dilogarithm}
 \label{sec:algorithm}
@@
 
-\section{Conclusions}
+\section{Summary}
 
@@
 the real dilogarithm have been compared w.r.t.\ their potential for
-the use of instruction\-/level parallelism.  A particular
-implementation of a rational minimax approximant has been proposed and
-compared to several existing implementations w.r.t.\ runtime.  A C
-implementations of the presented algorithm can be found in the
-appendix.  C++ and FORTRAN implementations can be found in the
-Polylogarithm library \cite{polylogarithm}.  Implementations in Julia
-and Rust can be found in \cite{PolyLog.jl} and \cite{polylog},
-respectively.
+the use of instruction\-/level (ILP) parallelism.  We found that a
+rational minimax approximant in combination with Estrin's scheme to
+evaluate the numerator/denominator polynomials provides both a
+sufficient numerical precision and a high degree of ILP, potentially
+leading to a reduced runtime.  A particular implementation of a
+rational minimax approximant has been proposed and compared to several
+existing implementations w.r.t.\ runtime.  A C implementations of the
+presented algorithm can be found in the appendix.  C++ and FORTRAN
+implementations can be found in the Polylogarithm library
+\cite{polylogarithm}.  Implementations in Julia and Rust can be found
+in \cite{PolyLog.jl} and \cite{polylog}, respectively.
 

### 2021-12-20 paper.tex
refine summary

@@
 sufficient numerical precision and a high degree of ILP, potentially
-leading to a reduced runtime.  A particular implementation of a
-rational minimax approximant has been proposed and compared to several
-existing implementations w.r.t.\ runtime.  A C implementations of the
-presented algorithm can be found in the appendix.  C++ and FORTRAN
+leading to a reduced runtime.  An approximant and a corresponding C
+implementation have been proposed and compared to several existing
+implementations w.r.t.\ their runtime.  The C implementation of the
+presented approximant can be found in the appendix.  C++ and FORTRAN
 implementations can be found in the Polylogarithm library

### 2021-12-20 paper.tex
avoid a non-breaking space

@@
 of the dilogarithm.  In the past many time-efficient algorithms have
-been invented (see e.g.~\cite{koelbigDilog,ginsberg,morris}) and many
+been invented (see e.g.\ \cite{koelbigDilog,ginsberg,morris}) and many
 are implemented, for example, in general-purpose physics and

### 2021-12-20 paper.tex
runtime -> run-time

@@
 parallelism (ILP) opens the opportunity for a further reduction of the
-runtime of the calculation of the dilogarithm.  To make use of ILP,
+run\-/time of the calculation of the dilogarithm.  To make use of ILP,
 the algorithm to calculate the dilogarithm must be formulated
@@
 calculation of the real dilogarithm $\Li$ is presented, which makes
-use of ILP, potentially resulting in a reduced runtime compared to
+use of ILP, potentially resulting in a reduced run\-/time compared to
 many algorithms presented so far.
@@
 floating\-/point multiplications, compared to the naive
-implementation, usually results in a shorter runtime.
+implementation, usually results in a shorter run\-/time.
 
-There are several different possibilities to reduce the runtime of
+There are several different possibilities to reduce the run\-/time of
 Horner's method by rewriting the series expansion of $\ln(1+x)$ such
@@
 
-\section{Runtime comparison}
+\section{Run\-/time comparison}
 
-In \tabref{tab:runtime} the runtime of different C implementations is
+In \tabref{tab:runtime} the run\-/time of different C implementations is
 compared to the implementation given in the appednix on two different
@@
 compiled with \texttt{g++} with optimization \texttt{-O2}.  The shown
-runtime denotes the total wall clock time for the calculation of
+run\-/time denotes the total wall clock time for the calculation of
 $\Li(x)$ for $10^6$ uniformly distributed random values $x\in[0,1/2]$.
@@
   \centering
-  \caption{Runtime in seconds for the calculation of $\Li(x)$ for
+  \caption{Run\-/time in seconds for the calculation of $\Li(x)$ for
     different implementations on different CPU architectures with
@@
 comparison) to use Horner's scheme to evaluate the numerator and the
-denominator.  Thus, the runtime difference of $\sim 30\%$ between
+denominator.  Thus, the run\-/time difference of $\sim 30\%$ between
 the two stems only from the fact that the implementation presented
@@
 by Cephes is by a factor two slower than the implementation prestend
-in this paper.  This longer runtime has several reasons: For one,
+in this paper.  This longer run\-/time has several reasons: For one,
 Cephes uses a rational function approximation with more terms in the
@@
 purely sequential.\footnote{For $x>1$ Cephes calculates $\ln(x)$
-  twice, which leads to a further increased runtime when a benchmark
+  twice, which leads to a further increased run\-/time when a benchmark
   interval with $x>1$ is chosen.}
@@
 scheme.  These algorithms thus do not make use of ILP, which is one
-reason for their longer runtime.  Furthermore, Algorithm 490 and most
+reason for their longer run\-/time.  Furthermore, Algorithm 490 and most
 series expansions in the GSL use floating\-/point divisions inside
-loops, which may be another reason for the increased runtime.
+loops, which may be another reason for the increased run\-/time.
 
@@
 Horner's scheme.  This is a reason why the implementation in
-\cite{root} has a larger runtime than the rational function
+\cite{root} has a larger run\-/time than the rational function
 approximants discussed above.
@@
 sufficient numerical precision and a high degree of ILP, potentially
-leading to a reduced runtime.  An approximant and a corresponding C
+leading to a reduced run\-/time.  An approximant and a corresponding C
 implementation have been proposed and compared to several existing
-implementations w.r.t.\ their runtime.  The C implementation of the
+implementations w.r.t.\ their run\-/time.  The C implementation of the
 presented approximant can be found in the appendix.  C++ and FORTRAN

### 2021-12-20 paper.tex
fix typos

@@
 
-As is well known, the performante of this naive implementation can be
+As is well known, the performance of this naive implementation can be
 improved by applying Horner's scheme, which reduces the number of
@@
 very similar to the results presented in Ref.~\cite{morris} (DILOG
-0011), although with slightly different coefficients.  In oder to
+0011), although with slightly different coefficients.  In order to
 allow for the use of ILP for the evaluation of the numerator and the
@@
 In \tabref{tab:runtime} the run\-/time of different C implementations is
-compared to the implementation given in the appednix on two different
-CPU architectures.  On each architecture the implentations are
+compared to the implementation given in the appendix on two different
+CPU architectures.  On each architecture the implementations are
 compiled with \texttt{g++} with optimization \texttt{-O2}.  The shown
@@
 function $\operatorname{Sp}(x)=\Li(1-x)$.  The implementation provided
-by Cephes is by a factor two slower than the implementation prestend
+by Cephes is by a factor two slower than the implementation presented
 in this paper.  This longer run\-/time has several reasons: For one,
@@
 implementation the polynomials are summed using Clenshaw's algorithm,
-which is also purely sequencial and slightly more complex than
+which is also purely sequential and slightly more complex than
 Horner's scheme.  This is a reason why the implementation in

### 2021-12-20 paper.tex
fix typos
language improvements

@@
 we will briefly show various possible implementations of this function
-and discuss the performance of the different implementations.  For
-brevity we will restrict ourselves to $x\in(-1,1]$.  In C the function
-in Eq.~\eqref{eq:taylor} could be implemented very naively as follows:
+and discuss their performance implications.  For brevity we will
+restrict ourselves to $x\in(-1,1]$.  In C the function in
+Eq.~\eqref{eq:taylor} could be implemented very naively as follows:
 %
@@
 the many (redundant) floating\-/point multiplications is usually so
-high that this formulated is in total very time\-/inefficient.
+high that this formulation is in total very time\-/inefficient.
 
@@
 %
-Note that the value $y=x^2$ can be re-used in the calculation of both
-the numerator and in the denominator.  This ``mixed'' implementation
-requires 8 floating\-/point multiplications, 1 floating\-/point
-division and 6 additions.  The rational function approximation
-\eqref{eq:mixed} may be faster for long numerator/denominator
-polynomials, where the cost of the floating\-/point division is
-smaller than the gain by the parallel execution of the terms in the
-numerator and denominator.
+Note, that the value $y=x^2$ can be re-used in the calculation of both
+the numerator and in the denominator.  A computation of $y^2$ is not
+necessary.  This ``mixed'' implementation requires 8 floating\-/point
+multiplications, 1 floating\-/point division and 6 additions.  The
+rational function approximation in Eq.~\eqref{eq:mixed} may be faster
+for long numerator/denominator polynomials, where the cost of the
+floating\-/point division is smaller than the gain by the parallel
+execution of the terms in the numerator and denominator.
 
@@
 \begin{align}
-  \Li(x) &= -\int_0^x \frac{\ln(1-t)}{t}dt, &
+  \Li(x) &= -\int_0^x \frac{\ln(1-t)}{t} \, \textnormal{d}t, &
   &x<1.
@@
 for example, using Horner's scheme or Clenshaw's algorithm
-\cite{clenshaw}.  These summation techniques are, however, purely
-sequential and thus do not make use of ILP.  In the following an
-approximation of the real dilogarithm is presented, which makes use of
-ILP, using a rational function approximation in combination with
-Estrin's scheme, similar to the approach used in
+\cite{clenshaw}, respectively.  These summation techniques are,
+however, purely sequential and thus do not make use of ILP.  In the
+following an approximation of the real dilogarithm is presented, which
+makes use of ILP, using a rational function approximation in
+combination with Estrin's scheme, similar to the approach used in
 Eq.~\eqref{eq:mixed}.
@@
 The approximant on the r.h.s.\ of Eq.~\eqref{eq:minimax} has a maximum
-relative error of around $5\cdot 10^{-17}$, suited for an evaluation
-of $\Li(x)$ with double precision.
+relative error of around $5\cdot 10^{-17}$.  This error is small
+enough so that the approximant Eq.~\eqref{eq:minimax} can be used for
+the evaluation of $\Li(x)$ with double precision.
 %
@@
 Note that the approximant on the r.h.s.\ of Eq.~\eqref{eq:minimax} is
-very similar to the results presented in Ref.~\cite{morris} (DILOG
+very similar to the approximant presented in Ref.~\cite{morris} (DILOG
 0011), although with slightly different coefficients.  In order to
 allow for the use of ILP for the evaluation of the numerator and the
-denominator we use Estrin's scheme,
+denominator we use Estrin's scheme and write
 %
 \begin{align}
-  \Li(x) &\approx x\frac{P(x)}{Q(x)}, \qquad 0\leq x \leq \frac{1}{2},
+  \Li(x) &\approx x\frac{P(x)}{Q(x)}, & &0\leq x \leq \frac{1}{2},
   \label{eq:final}
@@
 \begin{align}
-  P(x) &= p_0 + x p_1 + y (p_2 + x p_3) + z (p_4 + x p_5), \\
-  Q(x) &= q_0 + x q_1 + y (q_2 + x q_3) + z (q_4 + x q_5 + y q_6),
+  P(x) &= p_0 + p_1 x + (p_2 + p_3 x) y + (p_4 + p_5 x) z, \\
+  Q(x) &= q_0 + q_1 x + (q_2 + q_3 x) y + (q_4 + q_5 x + q_6 y) z,
 \end{align}
@@
 
-In \tabref{tab:runtime} the run\-/time of different C implementations is
-compared to the implementation given in the appendix on two different
-CPU architectures.  On each architecture the implementations are
-compiled with \texttt{g++} with optimization \texttt{-O2}.  The shown
-run\-/time denotes the total wall clock time for the calculation of
-$\Li(x)$ for $10^6$ uniformly distributed random values $x\in[0,1/2]$.
-The choice of this interval ensures that the measurement is only
-sensitive to the formulation of the approximant, but not to the
-transformation onto the interval $[0,1/2]$.%
+In \tabref{tab:runtime} the run\-/time of different C implementations
+for the real dilogarithm is compared to the implementation given in
+the appendix on two different CPU architectures.  On each architecture
+the implementations are compiled with \texttt{g++} with optimization
+\texttt{-O2}.  The shown run\-/time denotes the total wall clock time
+for the calculation of $\Li(x)$ for $10^6$ uniformly distributed
+random values $x\in[0,1/2]$.  The choice of this interval ensures that
+the measurement is only sensitive to the formulation of the
+approximant, but not to the transformation onto the interval
+$[0,1/2]$.%
 %
@@
 denominator.  Thus, the run\-/time difference of $\sim 30\%$ between
-the two stems only from the fact that the implementation presented
-in this paper uses Estrin's scheme to evaluate the numerator and the
+the two stems only from the fact that the implementation presented in
+this paper uses Estrin's scheme to evaluate the numerator and the
 denominator and as such can make use of ILP, while in the
 implementation of Ref.~\cite{morris} the numerator and the denominator
-must be evaluated sequentially.
+are evaluated sequentially.
 
@@
 scheme.  These algorithms thus do not make use of ILP, which is one
-reason for their longer run\-/time.  Furthermore, Algorithm 490 and most
-series expansions in the GSL use floating\-/point divisions inside
-loops, which may be another reason for the increased run\-/time.
+reason for their longer run\-/time.  Furthermore, Algorithm 490 and
+most series expansions in the GSL use expensive floating\-/point
+divisions inside loops, which may be another reason for the increased
+run\-/time.
 
-A representation in terms of Chebyshev polynomials \cite{luke} is
+The representation in terms of Chebyshev polynomials \cite{luke} is
 implemented in the ROOT program library \cite{root}.  In this
@@
 sufficient numerical precision and a high degree of ILP, potentially
-leading to a reduced run\-/time.  An approximant and a corresponding C
+leading to a reduced run\-/time.  A corresponding approximant and a C
 implementation have been proposed and compared to several existing

### 2021-12-20 paper.tex
improve formulation

@@
 corrections in quantum field theory.  The investigation of new models
-beyond the Standard Model of particle physics with a large parameter
-space require the evaluation of the dilogarithm (and other functions)
-billions of times.  Such investigations thus require a fast evaluation
-of the dilogarithm.  In the past many time-efficient algorithms have
-been invented (see e.g.\ \cite{koelbigDilog,ginsberg,morris}) and many
-are implemented, for example, in general-purpose physics and
-mathematics program libraries \cite{root,gsl,cephes}.
+beyond the Standard Model of particle physics with large parameter
+spaces require the numerical evaluation of the dilogarithm (and other
+functions) billions of times.  Such investigations thus require a fast
+evaluation of the dilogarithm.  In the past many time-efficient
+algorithms have been invented (see e.g.\
+\cite{koelbigDilog,ginsberg,morris,luke}) and many are implemented,
+for example, in general-purpose physics and mathematics program
+libraries \cite{root,gsl,cephes}.
 

### 2021-12-20 paper.tex
improve formulation

@@
 use formulations for the approximation of the dilogarithm with a
-strictly sequential execution of floating\-/point operations.  The
-development of modern CPUs with support for instruction\-/level
+strictly sequential execution of floating\-/point operations.  Here
+the development of modern CPUs with support for instruction\-/level
 parallelism (ILP) opens the opportunity for a further reduction of the

### 2021-12-21 paper.tex
use biblatex for nicer reference list
adding remark

@@
 
-\usepackage[utf8x]{inputenc}
+\usepackage[utf8]{inputenc}
 \usepackage[T1]{fontenc}
@@
 \usepackage{multirow}
+\usepackage[sorting=none]{biblatex}
+\addbibresource{paper.bib}
+
 \setkomafont{disposition}{\bfseries}
@@
 approximant, but not to the transformation onto the interval
-$[0,1/2]$.%
-%
+$[0,1/2]$.  Note that in applications where
+$x\in\mathbb{R}\backslash [0,1/2]$, the transformation onto the
+interval $[0,1/2]$ can have a significant impact on the run\-/time.
+
 \begin{table}[tb]
@@
 
-\bibliographystyle{ieeetr}
-\bibliography{paper}
+\printbibliography
 

### 2021-12-22 paper.tex
improve wording

@@
 \usepackage{multirow}
-\usepackage[sorting=none]{biblatex}
+\usepackage[sorting=none,citestyle=numeric-comp]{biblatex}
 \addbibresource{paper.bib}
@@
 evaluation of the dilogarithm.  In the past many time-efficient
-algorithms have been invented (see e.g.\
+algorithms have been developed (see e.g.\
 \cite{koelbigDilog,ginsberg,morris,luke}) and many are implemented,
@@
 the algorithm to calculate the dilogarithm must be formulated
-appropriately in order to allow for the parallel execution of multiple
-floating\-/point operations.  In this paper a method for the
+appropriately in order to allow for the simultaneous execution of
+multiple floating\-/point operations.  In this paper a method for the
 calculation of the real dilogarithm $\Li$ is presented, which makes
-use of ILP, potentially resulting in a reduced run\-/time compared to
-many algorithms presented so far.
+use of ILP, resulting in a reduced run\-/time compared to many
+algorithms presented so far, when executed on appropriate CPUs.
 
@@
 
-As a toy example we consider the function $\ln(1+x)$, which has the
-Taylor expansion
+As a toy example we consider the real function $\ln(1+x)$ for $x>-1$,
+which has the Taylor expansion
 %
@@
 Note, that the value $y=x^2$ can be re-used in the calculation of both
-the numerator and in the denominator.  A computation of $y^2$ is not
+the numerator and the denominator.  A computation of $y^2$ is not
 necessary.  This ``mixed'' implementation requires 8 floating\-/point
 multiplications, 1 floating\-/point division and 6 additions.  The
-rational function approximation in Eq.~\eqref{eq:mixed} may be faster
-for long numerator/denominator polynomials, where the cost of the
-floating\-/point division is smaller than the gain by the parallel
-execution of the terms in the numerator and denominator.
+rational function approximation in Eq.~\eqref{eq:mixed} may be
+beneficial for long numerator/denominator polynomials, where the cost
+of the floating\-/point division is smaller than the gain by the
+parallel execution of the terms in the numerator and denominator.
 
@@
 In the following an ILP-aware implementation of the real dilogarithm
-is presented, which allows for the parallel execution of different
+is presented, which allows for the simultaneous execution of multiple
 floating\-/point operations.  The real dilogarithm is defined as
@@
 however, purely sequential and thus do not make use of ILP.  In the
-following an approximation of the real dilogarithm is presented, which
+following an approximant of the real dilogarithm is presented, which
 makes use of ILP, using a rational function approximation in
@@
 0011), although with slightly different coefficients.  In order to
-allow for the use of ILP for the evaluation of the numerator and the
-denominator we use Estrin's scheme and write
+allow for the use of ILP in the evaluation of the numerator and the
+denominator on the r.h.s.\ of Eq.~\eqref{eq:minimax} we use Estrin's
+scheme and write
 %
@@
 the implementations are compiled with \texttt{g++} with optimization
-\texttt{-O2}.  The shown run\-/time denotes the total wall clock time
-for the calculation of $\Li(x)$ for $10^6$ uniformly distributed
+level \texttt{-O2}.  The shown run\-/time denotes the total wall clock
+time for the calculation of $\Li(x)$ for $10^6$ uniformly distributed
 random values $x\in[0,1/2]$.  The choice of this interval ensures that
@@
 purely sequential.\footnote{For $x>1$ Cephes calculates $\ln(x)$
-  twice, which leads to a further increased run\-/time when a benchmark
-  interval with $x>1$ is chosen.}
+  twice, which leads to a further increased run\-/time when a
+  benchmark interval with $x>1$ would be chosen.}
 
@@
 sufficient numerical precision and a high degree of ILP, potentially
-leading to a reduced run\-/time.  A corresponding approximant and a C
-implementation have been proposed and compared to several existing
-implementations w.r.t.\ their run\-/time.  The C implementation of the
-presented approximant can be found in the appendix.  C++ and FORTRAN
-implementations can be found in the Polylogarithm library
-\cite{polylogarithm}.  Implementations in Julia and Rust can be found
-in \cite{PolyLog.jl} and \cite{polylog}, respectively.
+leading to a reduced run\-/time when executed on an appropriate CPU.
+A corresponding approximant and a C implementation have been proposed
+and compared to several existing implementations w.r.t.\ their
+run\-/time.  The C implementation of the presented approximant can be
+found in the appendix.  C++, Fortran, Julia and Rust implementations
+can be found in \cite{polylogarithm,PolyLog.jl,polylog}, respectively.
 

### 2021-12-22 paper.tex
adding citation

@@
 \usepackage{multirow}
-\usepackage[sorting=none,citestyle=numeric-comp]{biblatex}
+\usepackage[sorting=none,citestyle=numeric-comp,eprint=false,isbn=false]{biblatex}
 \addbibresource{paper.bib}
@@
 As is well known, the performance of this naive implementation can be
-improved by applying Horner's scheme, which reduces the number of
-expensive floating\-/point multiplications to 7 and requires 6
-additions:
+improved by applying Horner's scheme \cite{horner}, which reduces the
+number of expensive floating\-/point multiplications to 7 and requires
+6 additions:
 %

### 2021-12-22 paper.tex
comment out footnote

@@
 numerator and denominator polynomials using Horner's scheme, which is
-purely sequential.\footnote{For $x>1$ Cephes calculates $\ln(x)$
-  twice, which leads to a further increased run\-/time when a
-  benchmark interval with $x>1$ would be chosen.}
+purely sequential.%
+% \footnote{For $x>1$ Cephes calculates $\ln(x)$ twice, which leads to a
+%   further increased run\-/time when a benchmark interval with $x>1$
+%   would be chosen.}
 

### 2021-12-27 paper.tex
fix typos

@@
 order.  However, the significant reduction of the number of
-floating\-/point multiplications, compared to the naive
-implementation, usually results in a shorter run\-/time.
+floating\-/point multiplications, compared to the naive implementation
+from above, usually results in a shorter run\-/time.
 
@@
 \begin{split}
-  \ln(1+x) \approx{}& x + \left(\frac{x}{3}-\frac{1}{2}\right)
-    y \\ &+\left[\left(\frac{x}{7}-\frac{1}{6}\right)
-    y+\frac{x}{5}-\frac{1}{4}\right] z,
+  \ln(1+x) \approx{}& x + y \left(\frac{x}{3}-\frac{1}{2}\right)
+    \\ &+ z \left[\frac{x}{5} - \frac{1}{4}
+      + y \left(\frac{x}{7}-\frac{1}{6}\right)
+    \right],
 \end{split}\label{eq:estrin}%
@@
 
-To exploit ILP further, one could use a rational function
+Another option to exploit ILP is to use a rational function
 approximation, such as a Padé approximant \cite{pade}, where the
@@
 approximant of $\ln(1+x)$ at the same order as Eq.~\eqref{eq:taylor}
-reads
+can be written as
 %
@@
   \ln(1+x) &\approx x
-  \frac{\frac{320 x^3}{40143}+\frac{44320 x^2}{120429}+\frac{51064
-    x}{40143}+1}{\frac{5120 x^3}{40143}+\frac{12320 x^2}{13381}+\frac{23712
-    x}{13381}+1}
+  \frac{
+    1 + \frac{51064}{40143} x + \frac{44320}{120429} x^2 + \frac{320}{40143} x^3
+  }{
+    1 + \frac{23712}{13381} x + \frac{12320}{13381} x^2 + \frac{5120}{40143} x^3
+  }
   \label{eq:pade}
@@
   &= x
-  \frac{\left(\frac{320 x}{40143}+\frac{44320}{120429}\right)y+\frac{51064
-    x}{40143}+1}{\left(\frac{5120 x}{40143}+\frac{12320}{13381}\right)y+\frac{23712
-    x}{13381}+1},
+  \frac{
+    1 + \frac{51064}{40143} x + y \left(\frac{44320}{120429} + \frac{320}{40143} x\right)
+  }{
+    1 + \frac{23712}{13381} x + y \left(\frac{12320}{13381} + \frac{5120}{40143} x\right)
+  },
   \label{eq:mixed}
@@
 denominator have been re-written using Estrin's scheme, allowing for
-further ILP.  A C implementation of Eq.~\eqref{eq:mixed} could read:
+further ILP, compared to Eq.~\eqref{eq:pade}.  A C implementation of
+Eq.~\eqref{eq:mixed} could read:
 %

### 2021-12-27 src/log1p_estrin.c

@@ -1,6 +1,6 @@
 double log1p_estrin(double x) {
    const double y = x*x;
    const double z = y*y;
-   return x + (1./3*x - 1./2)*y +
-      (y*(-1./6 + 1./7*x) + 1./5*x - 1./4)*z;
+   return x + y*(1./3*x - 1./2) +
+      z*(1./5*x - 1./4 + y*(-1./6 + 1./7*x));
 }

### 2021-12-27 paper.tex
refine text

@@
 necessary.  This ``mixed'' implementation requires 8 floating\-/point
-multiplications, 1 floating\-/point division and 6 additions.  The
-rational function approximation in Eq.~\eqref{eq:mixed} may be
-beneficial for long numerator/denominator polynomials, where the cost
-of the floating\-/point division is smaller than the gain by the
-parallel execution of the terms in the numerator and denominator.
+multiplications, 1 floating\-/point division and 6 additions.  A more
+optimized minimax rational function approximant, similar to
+Eq.~\eqref{eq:mixed}, but with a smaller error, may be preferable over
+a Taylor expansion when approximating a function over a larger
+interval.  At the same time such a rational function approximant
+naturally allows for ILP, as the numerator and denominator can be
+computed in parallel.  It is in particular beneficial for long
+numerator/denominator polynomials, where the cost of the
+floating\-/point division is smaller than the gain by the parallel
+computation of the terms in the numerator and denominator.
 

### 2021-12-27 paper.tex
improve wording

@@
 In the following an ILP-aware implementation of the real dilogarithm
-is presented, which allows for the simultaneous execution of multiple
-floating\-/point operations.  The real dilogarithm is defined as
+with double precision is presented, which allows for the simultaneous
+execution of multiple floating\-/point operations.  The real
+dilogarithm is defined as
 %
@@
 sufficient numerical precision and a high degree of ILP, potentially
-leading to a reduced run\-/time when executed on an appropriate CPU.
+leading to a reduced run\-/time when executed on a appropriate CPUs.
 A corresponding approximant and a C implementation have been proposed

### 2021-12-27 paper.tex
change section title
allow equation breaking
typo

@@
 \setkomafont{disposition}{\bfseries}
+\allowdisplaybreaks
 
@@
 naturally allows for ILP, as the numerator and denominator can be
-computed in parallel.  It is in particular beneficial for long
+computed in parallel.  This is in particular beneficial for long
 numerator/denominator polynomials, where the cost of the
@@
 
-\section{ILP-aware approximation of the real dilogarithm}
+\section{ILP-aware approximant of the real dilogarithm}
 \label{sec:algorithm}

### 2021-12-30 anc/Li2.c
reuse computed values

@@ -43,14 +43,14 @@ double li2(double x)
       s = 1;
    } else if (x < 1) {
       y = 1 - x;
-      r = PI*PI/6 - log(x)*log(1 - x);
+      r = PI*PI/6 - log(x)*log(y);
       s = -1;
    } else if (x == 1) {
       return PI*PI/6;
    } else if (x < 2) {
       const double l = log(x);
       y = 1 - 1/x;
-      r = PI*PI/6 - l*(log(1 - 1/x) + 0.5*l);
+      r = PI*PI/6 - l*(log(y) + 0.5*l);
       s = 1;
    } else {
       const double l = log(x);

### 2021-12-30 anc/Li2.c
consistent spacing of *

@@ -61,12 +61,12 @@ double li2(double x)
 
    const double y2 = y*y;
    const double y4 = y2*y2;
-   const double p = P[0] + y * P[1]
-      + y2 * (P[2] + y * P[3])
-      + y4 * (P[4] + y * P[5]);
-   const double q = Q[0] + y * Q[1]
-      + y2 * (Q[2] + y * Q[3])
-      + y4 * (Q[4] + y * Q[5] + y2 * Q[6]);
+   const double p = P[0] + y*P[1]
+      + y2*(P[2] + y*P[3])
+      + y4*(P[4] + y*P[5]);
+   const double q = Q[0] + y*Q[1]
+      + y2*(Q[2] + y*Q[3])
+      + y4*(Q[4] + y*Q[5] + y2*Q[6]);
 
    return r + s*y*p/q;
 }

### 2022-01-01 paper.tex
typo

@@
 the real dilogarithm have been compared w.r.t.\ their potential for
-the use of instruction\-/level (ILP) parallelism.  We found that a
+the use of instruction\-/level parallelism (ILP).  We found that a
 rational minimax approximant in combination with Estrin's scheme to

### 2022-01-05 paper.tex
last refinements

@@
 %
-Note, that the formulation using Horner's scheme requires the
+Note that the formulation using Horner's scheme requires the
 evaluation of the floating\-/point operations in strictly sequential
@@
 Horner's method by rewriting the series expansion of $\ln(1+x)$ such
-that more terms can be calculated in parallel, while the number
+that more terms can be calculated in parallel, while the number of
 floating\-/point operations is still kept small.  One possibility is
@@
 multiplications and 6 additions, i.e.\ one more multiplication than
-Horner's scheme.  However, in Eq.~\eqref{eq:split} both sums can then
+Horner's scheme.  However, in Eq.~\eqref{eq:split} both sums can
 be calculated in parallel, leading to a potential speed-up if the cost
@@
 where $y=x^2$.  In Eq.~\eqref{eq:mixed} the numerator and the
-denominator have been re-written using Estrin's scheme, allowing for
-further ILP, compared to Eq.~\eqref{eq:pade}.  A C implementation of
-Eq.~\eqref{eq:mixed} could read:
+denominator have been re-written using Estrin's scheme to reduce the
+number of floating\-/point multiplications compared to
+Eq.~\eqref{eq:pade}, while still allowing for some degree of ILP at
+the same time.  A C implementation of Eq.~\eqref{eq:mixed} could read:
 %
@@
 %
-Note, that the value $y=x^2$ can be re-used in the calculation of both
+Note that the value $y=x^2$ can be re-used in the calculation of both
 the numerator and the denominator.  A computation of $y^2$ is not
 necessary.  This ``mixed'' implementation requires 8 floating\-/point
-multiplications, 1 floating\-/point division and 6 additions.  A more
-optimized minimax rational function approximant, similar to
-Eq.~\eqref{eq:mixed}, but with a smaller error, may be preferable over
-a Taylor expansion when approximating a function over a larger
-interval.  At the same time such a rational function approximant
+multiplications, 1 floating\-/point division and 6 additions.  Note
+also that a rational function approximant such as Eq.~\eqref{eq:mixed}
 naturally allows for ILP, as the numerator and denominator can be
@@
 floating\-/point division is smaller than the gain by the parallel
-computation of the terms in the numerator and denominator.
+computation of the terms in the numerator and denominator.  In order
+to further reduce the error and/or the number of terms in the
+approximant, a more ``optimized'' minimax rational function
+approximant may be used instead of a Taylor expansion or Padé
+approximant when approximating a function over a larger interval.
 
@@
 terms of Chebyshev polynomials of the first kind $T_n(x)$ is given in
-\cite{luke},
+Ref.~\cite{luke},
 %
@@
 comparison) to use Horner's scheme to evaluate the numerator and the
-denominator.  Thus, the run\-/time difference of $\sim 30\%$ between
+denominator.  Thus, the run\-/time difference of around $30\%$ between
 the two stems only from the fact that the implementation presented in
@@
 leading to a reduced run\-/time when executed on a appropriate CPUs.
-A corresponding approximant and a C implementation have been proposed
+A corresponding approximant and a C implementation have been presented
 and compared to several existing implementations w.r.t.\ their
